---
title: "Fitting the Memory Measurement Model ($M^3$) with the bmm R Package: A Tutorial"
shorttitle: "Fitting the $M^3$ with the bmm R Package"
author:
  - name: Gidon T. Frischkorn
    orcid: 0000-0002-5055-9764
    email: gidon.frischkorn@psychologie.uzh.ch
    corresponding: true
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Isabel Courage
    orcid: 0009-0000-8493-2240
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Chenyu Li
    orcid: 0000-0001-8815-6189
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
      - name: University of Chicago
        department: Department of Psychology
author-note:
  disclosures:
    study-registration: "This study was not preregistered."
    data-sharing: "Data, analysis code, and companion R scripts for all three tutorials are available at [repository link]."
    conflict-of-interest: "The authors declare no conflicts of interest."
    financial-support: "[FUNDING INFORMATION]"
    gratitude: "[ACKNOWLEDGMENTS]"
    authorship-agreements: |
      Gidon T. Frischkorn: Conceptualization, Methodology, Software, Formal Analysis, Visualization, Writing -- Original Draft.
      Isabel Courage: Software, Formal Analysis, Writing -- Review & Editing.
      Chenyu Li: Software, Data Curation, Writing -- Review & Editing.
format:
  apaquarto-pdf:
    keep-tex: true
  apaquarto-docx: default
floatsintext: true
bibliography: references.bib
abstract: |
  The memory measurement model ($M^3$) decomposes categorical response data from memory tasks into latent activation parameters, separating item memory from context-specific binding via competitive retrieval. Fitting $M^3$ has so far required custom JAGS or Stan code for each task design. The bmm R package provides a standardized interface to specify, fit, and evaluate Bayesian hierarchical $M^3$ with minimal code. This tutorial demonstrates the bmm workflow through three worked examples: fitting a simple span $M^3$ to cued recall data, where Bayes factor model comparison favors the softmax over the simple choice rule; extending the workflow to a complex span task with distractor categories and a filtering parameter; and defining a fully custom $M^3$ for a memory updating task, validating it through parameter recovery across a range of sample sizes and trial counts. Complete R scripts accompany each tutorial. After completing the tutorials, readers will be able to fit standard and custom $M^3$ to their own data and validate new model specifications through simulation.
keywords:
  - working memory
  - measurement model
  - multinomial model
  - Bayesian estimation
  - R package
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
#| eval: true
pacman::p_load(here, bmm, brms, tidyverse, tidybayes, patchwork)
source(here("functions", "clean_plot.R"))

# ---- Load fitted models for inline reporting ----
fit_m3_ss_softmax <- readRDS(here("output", "fit_m3_ss_softmax.rds"))
fit_m3_ss_simple  <- readRDS(here("output", "fit_m3_ss_simple.rds"))
fit_m3_cs         <- readRDS(here("output", "fit_m3_cs.rds"))
bridge_softmax    <- readRDS(here("output", "bridge_m3_ss_softmax.rds"))
bridge_simple     <- readRDS(here("output", "bridge_m3_ss_simple.rds"))

# ---- Convergence diagnostics across all models ----
get_diagnostics <- function(fit) {
  s <- summary(fit)
  rhats <- s$fixed[, "Rhat"]
  ess   <- s$fixed[, "Bulk_ESS"]
  for (g in names(s$random)) {
    rhats <- c(rhats, s$random[[g]][, "Rhat"])
    ess   <- c(ess,   s$random[[g]][, "Bulk_ESS"])
  }
  list(rhat = max(rhats, na.rm = TRUE), ess = min(ess, na.rm = TRUE))
}

diag_sm <- get_diagnostics(fit_m3_ss_softmax)
diag_si <- get_diagnostics(fit_m3_ss_simple)
diag_cs <- get_diagnostics(fit_m3_cs)

val_rhat_all <- sprintf("%.3f", max(diag_sm$rhat, diag_si$rhat, diag_cs$rhat))
val_ess_all  <- format(round(min(diag_sm$ess, diag_si$ess, diag_cs$ess)),
                       big.mark = ",")
val_rhat_t2  <- sprintf("%.3f", diag_cs$rhat)
val_ess_t2   <- format(round(diag_cs$ess), big.mark = ",")

# ---- Bayes factor: softmax vs simple ----
bf_result   <- bridgesampling::bf(bridge_softmax, bridge_simple)
bf10_choice <- bf_result$bf

bf10_med <- median(bf10_choice)
bf10_fmt <- if (bf10_med >= 10000) {
  ev <- floor(log10(bf10_med))
  cv <- bf10_med / 10^ev
  sprintf("$%.1f \\times 10^{%d}$", cv, ev)
} else {
  sprintf("%.1f", bf10_med)
}

# ---- Evidence ratios ----
fmt_er <- function(x) {
  if (is.na(x)) return("NA")
  if (is.infinite(x) || x >= 10000) return("> 10,000")
  if (x >= 1000) return(format(round(x), big.mark = ","))
  if (x >= 10) return(sprintf("%.1f", x))
  sprintf("%.2f", x)
}

h_ca      <- hypothesis(fit_m3_ss_softmax, "c_expopenset = a_expopenset")
er_ca_fmt <- fmt_er(1/h_ca$hypothesis$Evid.Ratio)

h_ca_closed      <- hypothesis(fit_m3_ss_softmax, "c_expclosedset = a_expclosedset")
er_ca_closed_fmt <- fmt_er(1/h_ca_closed$hypothesis$Evid.Ratio)

h_ss_c      <- hypothesis(fit_m3_ss_softmax,
  "(c_expopenset:ss_lin + c_expclosedset:ss_lin) / 2 = 0")
er_ss_c_fmt <- fmt_er(1/h_ss_c$hypothesis$Evid.Ratio)

h_ss_a      <- hypothesis(fit_m3_ss_softmax,
  "(a_expopenset:ss_lin + a_expclosedset:ss_lin) / 2 = 0")
er_ss_a_fmt <- fmt_er(1/h_ss_a$hypothesis$Evid.Ratio)

h_exp_ss_c      <- hypothesis(fit_m3_ss_softmax,
  "c_expopenset:ss_lin = c_expclosedset:ss_lin")
er_exp_ss_c_fmt <- fmt_er(h_exp_ss_c$hypothesis$Evid.Ratio)

h_f      <- hypothesis(fit_m3_cs, "f_conditionpre - f_conditionretro = 0")
er_f_fmt <- fmt_er(h_f$hypothesis$Evid.Ratio)

h_a_dist      <- hypothesis(fit_m3_cs,
  "a_conditioncontrol - (a_conditionpre + a_conditionretro) / 2 = 0")
er_a_dist_fmt <- fmt_er(1/h_a_dist$hypothesis$Evid.Ratio)

h_c_dist      <- hypothesis(fit_m3_cs,
  "c_conditioncontrol - (c_conditionpre + c_conditionretro) / 2 = 0")
er_c_dist_fmt <- fmt_er(1/h_c_dist$hypothesis$Evid.Ratio)

# ---- Tutorial 2: f on probability scale ----
draws_cs <- as_draws_df(fit_m3_cs) |> as_tibble()

f_pre_hdci <- tibble(f = plogis(draws_cs$b_f_conditionpre)) |>
  mean_hdci(f, .width = 0.95)
f_retro_hdci <- tibble(f = plogis(draws_cs$b_f_conditionretro)) |>
  mean_hdci(f, .width = 0.95)

val_f_pre_m    <- sprintf("%.2f", f_pre_hdci$f)
val_f_pre_lo   <- sprintf("%.2f", f_pre_hdci$.lower)
val_f_pre_hi   <- sprintf("%.2f", f_pre_hdci$.upper)
val_f_retro_m  <- sprintf("%.2f", f_retro_hdci$f)
val_f_retro_lo <- sprintf("%.2f", f_retro_hdci$.lower)
val_f_retro_hi <- sprintf("%.2f", f_retro_hdci$.upper)

# ---- Tutorial 3: Recovery correlations ----
# Reproduce true generating parameters using the same seed as the script
set.seed(2025)
N_sim <- 50
true_pars_t3 <- tibble(
  participant = 1:N_sim,
  c_link = rnorm(N_sim, 2.0, 0.50),
  a_link = rnorm(N_sim, 1.0, 0.50),
  d_link = rnorm(N_sim, 0.85, 0.65),
  e_link = rnorm(N_sim, 0.00, 0.35),
  r_link = rnorm(N_sim, -0.69, 0.38)
)

t3_file <- here("output", "fit_m3_custom_simple_N50_T25.rds")
if (file.exists(t3_file)) {
  fit_t3 <- readRDS(t3_file)
  co_t3  <- coef(fit_t3)
  par_names_t3 <- c("a", "c", "d", "e", "r")
  indiv_corrs <- sapply(par_names_t3, function(p) {
    est <- co_t3$participant[, , paste0(p, "_Intercept")]
    cor(true_pars_t3[[paste0(p, "_link")]], est[, "Estimate"])
  })
} else {
  indiv_corrs <- setNames(rep(NA_real_, 5), c("a", "c", "d", "e", "r"))
}

val_cor_a <- sprintf("%.2f", indiv_corrs["a"])
val_cor_c <- sprintf("%.2f", indiv_corrs["c"])
val_cor_d <- sprintf("%.2f", indiv_corrs["d"])
val_cor_e <- sprintf("%.2f", indiv_corrs["e"])
val_cor_r <- sprintf("%.2f", indiv_corrs["r"])
```

# Introduction

Observed performance in memory and decision tasks is often summarized by overall accuracy. Yet, for many tasks accuracy collapses across qualitatively distinct response categories and error types that may arise from different latent processes. In particular, different error types are diagnostically informative because different cognitive processes imply different conditional response probabilities over observable outcomes. Multinomial processing tree (MPT) models formalize this logic by treating categorical responses — including distinct error classes — as consequences of latent states such as successful retrieval, guessing, or source confusion [@batchelder1999theoretical; @erdfelder2009multinomial]. Similarly, process-dissociation approaches demonstrate that patterns of correct and incorrect responses across task constraints can separate the contributions of multiple processes within a single task [@jacoby1991process; @yonelinas2012process]. In sum, the full multinomial response distribution - not accuracy alone - carries more information about latent cognitive mechanisms than a single accuracy score.

Working memory research frequently relies on tasks that inherently yield response frequencies across discrete outcome categories, such as correct responses, feature confusions, location errors, or intrusions [@oberauer2018benchmarks]. For discrete feature spaces (e.g., words, numbers, letters), these data are naturally multinomial: each retrieval produces one of several mutually exclusive response types. Thus, theoretical claims can often be evaluated by how experimental manipulations and individual differences redistribute probability mass across these categories.

The recently proposed memory measurement model [$M^3$; @oberauer2019simple] framework provides one such decomposition. Its core assumptions are that 1) multiple activation sources contribute to each response category and that  2) selection among categories follows a competitive retrieval rule [@luce1959individual]. Although the $M^3$ framework was introduced for working memory tasks, the general architecture is domain-general: whenever a task yields categorical responses that can be traced to distinct latent activation sources driving competitive selection, the framework applies.[^1]

[^1]: Readers familiar with multinomial processing tree (MPT) models will note similarities: both frameworks decompose categorical response distributions into latent process parameters. However, whereas MPT models use discrete processing trees with binary branching probabilities [@batchelder1999theoretical; e.g., @heck2018treebugs for Bayesian hierarchical implementations], $M^3$ uses continuous activation values with competitive selection, making it particularly suited for theories emphasizing graded memory strength [for a detailed comparison, see @oberauer2019simple].

Because the $M^3$ framework is flexible — @oberauer2019simple already introduced more than five model architectures — the activation formulas that drive response selection typically vary across tasks and response categories. Each new application therefore required translating its model into custom JAGS or Stan code, making implementations error-prone and hard to reuse. The `bmm` R package [@bmm2025] addresses this limitation by providing a standardized workflow to specify, fit, and evaluate Bayesian hierarchical $M^3$ with minimal code, while remaining flexible enough to accommodate new task designs and custom model specifications. In this tutorial, we illustrate this approach with three worked examples covering the standard simple span (ss) and complex span (cs) tasks, and a custom extension.

# The $M^3$ Framework

Consider a cued recall task (e.g., remembering letters in serial positions or icons in spatial positions) in which a participant sees a retrieval cue and selects a response from a set containing the correct target, other studied items, and new lures. Each response falls into one of three categories: *correct* — the item associated with the retrieval cue, *other-list intrusion* — a studied item associated with a different cue, or *not-presented lure* (NPL) — an item not presented during the trial. The $M^3$ framework explains each category's selection frequency by specifying activation formulas: different activation sources feed into each category, and response probabilities arise as the result of applying choice rules that implement competitive selection among all options.

## Activation Formulas and Model Parameters

Each response category's activation formula specifies which latent sources contribute to its selection probability. Because the $M^3$ framework is not committed to any particular memory theory, different assumptions about the processes limiting memory performance can be accommodated through different activation formulas. 

Each response category usually receives activation from one or more latent sources. For the cued recall task introduced above, the theoretically assumed activation sources are:

-   $b$: **baseline activation**, the tendency to select any response option in the absence of memory-based evidence. This could reflect guessing, response bias, or other non-mnemonic factors that contribute to selection probabilities across all categories.
-   $a$: **item activation**, or general memory strength for studied items. An item that was encountered during study receives this activation regardless of its position or context.
-   $c$: **context binding**, the strength of the association between an item and its specific study context (e.g., serial position, paired associate).

These activation sources or additional mechanisms are specified as parameters that can vary and are estimated from data. Using these parameters, we can specify how they are combined to produce total activations for each response category. The simple span model proposed by @oberauer2019simple assumes that these activation sources combine additively, with the correct target receiving activation from all three sources, other studied items receiving item activation but lacking context binding, and lures receiving only baseline activation:

$$
\begin{aligned}
\text{A}_{\text{correct}} &= b + a + c \\
\text{A}_{\text{other}} &= b + a \\
\text{A}_{\text{lures}} &= b
\end{aligned}
$$

Another perspective on the interplay of these processes could also assume that context activation boosts item activation, such that the correct target receives a multiplicative boost from context binding, while other studied items receive only item activation:

$$
\begin{aligned}
\text{A}_{\text{correct}} &= b + a \cdot (1 + c) \\
\text{A}_{\text{other}} &= b + a \\
\text{A}_{\text{lures}} &= b
\end{aligned}
$$

These two examples illustrate how distinct theoretical perspectives can be encoded through the activation formulas.[^2] The `bmm` package provides pre-specified formulas for the simple span and complex span versions proposed by @oberauer2019simple, and also allows users to define custom formulas for novel task designs.

[^2]: Although the $M^3$ framework was concieved as a measurement tool, it can thus be used to implement and test specific theoretical assumptions. For example, similarity between spatial positions could be implemented by other list items getting a partial boost from context binding dependent on their distance to the currently cued position. Thereby the $M^3$ framework sits at the intersection of measurement and process models, it can account for the interplay of different processes, but does not provide any implementation of the activation sources and why they might differ in strength.

## Competitive Selection and Choice Rules

To link activations to observed responses, each category's activation must be converted into a response probability. The $M^3$ framework assumes competitive selection, that is a category's probability depends on its activation relative to all competitors. For implementing any choice rules that implement competivie selection the number of all competitors, that is options per category, needs to be known. Therefore, tasks designed for $M^3$ should use a recognition format in which the experimenter controls the available response options instead of free-recall where participants generate their response set internally.

Given these option counts, $M^3$ provides two choice rules that differ in how activations are transformed before normalization. The *simple* (or linear) choice rule [@luce1959individual] normalizes activations directly:

$$P(k) = \frac{n_k \cdot \text{A}_k}{\sum_{j=1}^{K} n_j \cdot \text{A}_j}$$

where $\text{A}_k$ is the total activation for category $k$ and $n_k$ is the number of response options in that category. This rule requires all activations to be positive. To identify the scaling of activations, the baseline $b$ is fixed at a small positive value (0.1 by default).

The *softmax* (or exponential) choice rule applies an exponential transformation before normalization:

$$P(k) = \frac{n_k \cdot \exp(\text{A}_k)}{\sum_{j=1}^{K} n_j \cdot \exp(\text{A}_j)}$$

Because $\exp(x) > 0$ for all $x$, softmax guarantees valid probabilities regardless of activation values. Again, to identify the scaling of activations, the baseline is fixed at $b = 0$ (since $\exp(0) = 1$, lures receive a positive baseline contribution automatically).

The softmax rule is formally equivalent to an $n$-alternative forced-choice signal detection model with Gumbel-distributed noise [@thurstone1927law; @yellott1977relationship]. Under this interpretation, each category's activation represents a deterministic signal, independent Gumbel noise is added to each alternative, and the observer selects the category with the highest noisy value. This equivalence gives activation parameters a principled interpretation as discriminability measures. Recent empirical comparisons have favored the softmax rule over the simple rule for working memory data [@oberauer2023measurement; @li2026filtering] although the simple choice rule was endorsed initially [@oberauer2019simple]. Tutorial 1 compares both rules; Tutorials 2 and 3 use the softmax rule as the default.

## Model Versions

The `bmm` package provides three $M^3$ versions, two of which implement canonical paradigms often used in working memory research. The versions differ in the response categories and estimated parameters. Here we shortly introduce them, the following tutorials provide more details.

The **simple span** (`ss`) version implements $M^3$ for cued recall tasks separating three response categories (correct response, other, lures) with two free parameters ($a$, $c$). It implements the additive activation formula proposed by @oberauer2019simple.

The **complex span** (`cs`) version extends the simple span to tasks that interleave distractor processing with memorization. It requires that memoranda and distractors are drawn from the same material so that distractors can plausibly intrude during retrieval [for examples, see @oberauer2019simple; @li2026filtering]. The `cs` version separates five response categories (correct, paired distractor, other memorandum, other distractor, and NPL) with three free parameters ($a$, $c$, $f$). The additional parameter $f$ captures distractor filtering efficiency on a scale from 0 to 1: $f = 0$ indicates perfect filtering (distractors receive only baseline activation), while $f = 1$ indicates no filtering (distractors are activated as strongly as memoranda).

The **custom** version allows user-defined response categories, activation formulas, and parameters. This version allows researchers to specify $M^3$ of arbitrary complexity for novel task designs, as long as the model remains identified. The only requirement is that all activation formulas include the baseline activation $b$, which reflects random guessing if no other activations were included. The `bmm` package provides a flexible syntax for defining custom models, and Tutorial 3 demonstrates how to validate a custom model through parameter recovery simulations.

## The bmm Workflow

Fitting $M^3$ with the `bmm` R package [@bmm2025] follows a three-step workflow that is consistent for all measurement models implemented in the package:

1.  **Define the model**: The package includes a constructor function `m3()` that generates a model object with all relevant information for model specification. This entails the variable names of response categories in the data, the number of response options per category, the model version, and the choice rule.
2.  **Specify predictor formulas**: The `bmm` package provides a `bmmformula()` or shorthand `bmf()` function to specify the linear prediction formulas that define how experimental conditions and individual differences predict each activation parameter. The `bmmformula` syntax uses the same formula syntax as `brms` [@burkner2017brms] and also includes random effects and non-linear formulas.[^3]
3.  **Fit the model**: The model object and prediction formula are passed to `bmm()` together with the data. This function translates the specification into a `brms` model and passes it to Stan [@carpenter2017stan] for Bayesian estimation.

[^3]: For details on the `bmmformula` syntax, see <https://venpopov.com/bmm/articles/bmm_bmmformula.html>.

Because `bmm` builds on `brms`, fitted models inherit most of the `brms` post-processing ecosystem, including posterior predictive checks, hypothesis tests, and model comparison.[^4]

[^4]: We are continously working on extending the post-processing functions to accommodaate the specific requirements of the measurement models implemented in the `bmm` R package. 

# Tutorial 1: Simple Span $M^3$

This tutorial walks through the complete $M^3$ workflow for the *simple span* (ss) version using data from @oberauer2019bindings, who measured cued recall with words presented in boxes distributed horizontally on the screen. Two experiments (20 participants each) differed in stimulus set type: Experiment 1 used new words on each trial (open set), while Experiment 2 reused a small set of words across trials (closed set). Set size (2, 4, 6, 8 pairs) varied within participants. For more details on the experimental procedure please refer to the original publication. The full companion script available in the online supplement (`scripts/tutorial1_simple_span.R`) contains all code; we reproduce the key steps here.

## Data

Each recognition trial presents a retrieval cue ("?") in one of the boxes together with a set of response options. This set contained the correct target, other studied items from the same list, and not-presented lures. Participants selected one item, producing a categorical response: *correct*, *other* (intrusion from a different list position), or *NPL* (not-presented lure). The number of response options per category varied across 24 recognition conditions (combinations of set size, list items in the recognition set, and number of lures) that we will not model in detail here.

A $M^3$ in `bmm` expects aggregated data with one row per participant and condition. Each row must contain (a) response frequency counts for each response category — that is how often the participant chose each category — and (b) the number of response options available per category. For the simple span version, the three response categories are *correct*, *other*, and *NPL*, so the dataset requires three frequency columns (`corr`, `other`, `npl`) and three option-count columns (`n_corr`, `n_other`, `n_npl`). Any predictors of interest (here, mean-centered set size `ss_lin` and experiment `exp`) are included as additional columns. The companion script (`scripts/tutorial1_simple_span.R`) shows how to aggregate trial-level data from @oberauer2019bindings into this format; the resulting dataset has 960 rows (40 participants $\times$ 24 recognition conditions). Below are the first rows of the data set:

```{r}
#| label: t1-data-prep
#| eval: true
#| echo: false
# Read both experiments; create unique participant IDs
col_names <- c("id", "session", "block", "trial", "setsize",
               "rsizeList", "rsizeNPL", "tested", "response", "rcat", "rt")

read_exp <- function(file, exp_label) {
  read.table(here("data", file), header = FALSE,
             col.names = col_names, fill = TRUE) %>%
    drop_na() %>%
    mutate(exp = exp_label,
           id  = paste(id, exp_label, sep = "_"))
}

data_raw <- bind_rows(
  read_exp("Oberauer_2019_SimpleSpan_Exp1.dat", "openset"),
  read_exp("Oberauer_2019_SimpleSpan_Exp2.dat", "closedset")
)
```

```{r}
#| label: t1-data-agg
#| eval: true
#| echo: false
# Keep recognition trials only, aggregate to response frequencies
data_agg <- data_raw %>%
  filter(rsizeList + rsizeNPL > 0) %>%
  group_by(id, exp, setsize, rsizeList, rsizeNPL) %>%
  summarise(corr  = sum(rcat == 1),
            other = sum(rcat == 2),
            npl   = sum(rcat == 3), .groups = "drop") %>%
  mutate(n_corr  = 1,
         n_other = rsizeList - 1,  # rsizeList includes the correct item
         n_npl   = rsizeNPL,
         ss_lin  = setsize - mean(setsize),
         exp     = factor(exp))

head(data_agg)
```

## Model Specification

Specifying an $M^3$ requires two R objects: 1) the $M^3$ object using the  `m3()` and 2) a predictor formula (`bmf()`). For creating the $M^3$ object you need to call the `m3()` model constructor an pass a) the name of the variables that contain the response frequency in all response categories, b) the columns containing option counts or a named vector with the number of response options if they are constant for all observations, c) the model version, and d) the choice rule. If no model version is specified the `"custom"` version is used, and if no choice rule is specified the `"softmax"` rule is used as default.

When explicitly passing all arguments, the specification of the simple span model object for the data from @oberauer2019bindings looks like this:

```{r}
#| label: t1-model-spec
m3_model_ss <- m3(
  resp_cats   = c("corr", "other", "npl"),
  num_options = c("n_corr", "n_other", "n_npl"),
  version     = "ss",
  choice_rule = "softmax"
)
```

Once we have the model object specified, we can assess which parameters the model contains by calling `m3_model_ss$parameters`. For the simple span version, the parameters are `a` (item activation) and `c` (context binding). The baseline parameter `b` is fixed and not estimated as can be seen in `m3_model_ss$fixed_parameters`.

One important note applies to all pre-built `m3()` versions: activation formulas are assigned to response categories *by position*, not by name. The `resp_cats` vector must therefore follow the canonical order for the chosen version — for `ss`: `corr, other, npl`. Supplying categories in a different order will silently assign the wrong activation formulas to the wrong categories, producing incorrect estimates without any warning. The same positional rule applies to `num_options`.

With this information, we know that we can specify predictor formulas for the two activation parameters $a$ and $c$. Because the two experiments in @oberauer2019bindings are between-subjects, we use cell means coding (`0 + exp`) so that each experiment receives its own intercept rather than a difference from a reference level or grand mean. Similar to the original study by @oberauer2019bindings, we entered set size as a linear predictor (`exp:ss_lin`), and random effects allowed both intercepts and slopes to vary across participants, with separate variance components per experiment (`gr(id, by = exp)`)[^5].

[^5]: For details on the specification of the `bmmformula`, see the `bmm` package documentation and the documentation of the `brmsformula` function on which it builds.

```{r}
#| label: t1-formula
m3_formula <- bmf(
  c ~ 0 + exp + exp:ss_lin + (1 + ss_lin | gr(id, by = exp)),
  a ~ 0 + exp + exp:ss_lin + (1 + ss_lin | gr(id, by = exp))
)
```

When using the softmax rule, $a$ and $c$ are estimated on the identity link (native activation scale) which can be seen via `m3_model_ss$links`. Because there is some information about parameter ranges available from previous literature, `bmm` specifies weakly informative priors on the intercepts and effect parameters. This is different to `brms`, which uses flat priors for effects by default. These default priors can be inspected with `default_prior(m3_formula, m3_model_ss, data = data_agg)`. All these priors can be overridden if needed. Throughout this tutorial we use the defaults; Tutorial 2 shows how to set custom priors when needed.

## Model Fitting

The `bmm()` function combines the model definition and formula, and passes them together with the data to `brms` and Stan, and returns a fitted model object. All models in this tutorial were estimated with the default setting of 4 chains of 2,000 iterations each (1,000 warmup), using `cmdstanr` as the backend. We set `sample_prior = "yes"` so that prior samples are available for later hypothesis tests. In addition, `save_pars = save_pars(all = TRUE)` specifies that all model parameters and likelihood results should be saved, which is necessary to perfrom model comparisons via Bayes Factors. The `file` argument caches the fitted model to disk; re-running the script loads the cached fit rather than re-fitting.[^6] 

[^6]: These are all additional options that are passed to `brms` and can be modified as needed. For details on these additional arguments, please refer to `brm()` documentation.

```{r}
#| label: t1-fit
#| eval: false
fit_m3_ss_softmax <- bmm(
  formula      = m3_formula,
  model        = m3_model_ss,
  data         = data_agg,
  cores        = 4,
  backend      = "cmdstanr",
  sample_prior = "yes",
  save_pars    = save_pars(all = TRUE),
  file         = here("output", "fit_m3_ss_softmax")
)
```

As for all Bayesian models we should check convergence after fitting. The summary contains `summary(fit_m3_ss_softmax)` convergence statistics $\hat{R}$ for all major model parameters. All $\hat{R}$ values [@gelman1992inference] should be below 1.01 and effective sample sizes should be adequate. For all models in this tutorial, these criteria were met ($\hat{R} \leq$ `r val_rhat_all`, bulk ESS $>$ `r val_ess_all`).[^7] Ocassionally, Stan may report warnings for divergent transitions, hitting the maximum tree depth, or low effective sample sizes. These issues can often be resolved by increasing the number of iterations, adjusting the `adapt_delta` parameter, or reparameterizing the model. For guidance on addressing convergence issues, see the `brms` documentation and resources on Bayesian modeling best practices. In cases where more samples and tuning do not resolve convergence issues, it may be necessary to simplify the model or reconsider the model specification.

[^7]: In principle, the measurement models implemented in `bmm` are similar to all other Bayesian generalized linear models and thus similar criteria for convergence apply. For guidance on Bayesian convergence diagnostics, see @gabry2019visualization and the `brms` documentation.

## Model Evaluation

Before interpreting parameter estimates, we verify that the fitted model matches the observed data. A posterior predictive check compares observed response proportions to model-predicted proportions; systematic discrepancies would indicate that the activation formulas or choice rule fail to capture the data-generating process. We used `posterior_epred()` provided from the `brms` package to obtain expected counts per category for each observation. We then averaged across posterior draws, and aggregated to participant-level proportions by set size and experiment. @fig-t1-pp-check shows the result. Across all conditions and response categories, predicted proportions fall within one standard error of the observed means, indicating that the simple span $M^3$ with softmax choice rule adequately captures the response distributions.

```{r}
#| label: fig-t1-pp-check
#| eval: true
#| echo: false
#| fig-cap: "Posterior predictive check for the simple span $M^3$ (softmax choice rule). Points and lines show mean response proportions by set size and experiment; error bars show $\\pm 1$ SE across participants. Color distinguishes observed from predicted values."
#| fig-alt: "Line plot comparing observed and predicted response proportions across set sizes 2, 4, 6, and 8 for three response categories (correct, other, NPL) in open-set and closed-set experiments. Predicted proportions closely track observed values in all conditions."
knitr::include_graphics(here("figures", "tutorial1_pp_check_softmax.pdf"))
```

## Parameter Estimates

With the model validated, we can examine how item memory ($a$) and context binding ($c$) varied across set sizes and experiments. Under the softmax choice rule, both parameters are on the identity link, so the posterior draws are directly interpretable as activation values. To compute condition-specific estimates, we reconstructed the linear predictor from the fixed effects: for each set size $s$, $\hat{a}_{\text{exp}} = \beta_{a,\text{exp}} + \beta_{a,\text{exp:ss\_lin}} \cdot (s - \bar{s})$, and analogously for $c$. The companion script extracts draws using  `tidybayes::as_draws_df()` and computes 95% highest-density intervals [HDIs; @mcelreath2020statistical] via `tidybayes::mean_hdci()`.

@fig-t1-param-estimates shows the resulting estimates. Context binding ($c$) is descriptively higher than item memory ($a$) across all set sizes in both experiments, with the gap being especially pronounced in the closed-set experiment. Both parameters decrease with set size, but the decline is substantially steeper for $c$ than for $a$, consistent with binding being more sensitive to memory load. The two experiments differ markedly in item activation: $a$ is notably higher in the open-set experiment than in the closed-set experiment. This likely reflects cross-trial interference in the closed-set condition, where the same words recur across trials and become associated with multiple contexts, reducing effective item-specific memory strength. These resuls are consistent with the original findings by @oberauer2019bindings, who also found higher $a$ in the open-set experiment and a steeper decline of $c$ with set size. 

```{r}
#| label: fig-t1-param-estimates
#| eval: true
#| echo: false
#| fig-cap: "Posterior estimates of item memory ($a$) and context binding ($c$) across set sizes for the open-set and closed-set experiments (softmax choice rule). Points show posterior means; error bars show 95\\% HDIs."
#| fig-alt: "Point-range plot showing posterior means and 95% HDIs for item memory (a) and context binding (c) across set sizes 2, 4, 6, and 8, separately for open-set and closed-set experiments. Both parameters decrease with set size, with c consistently higher than a."
knitr::include_graphics(here("figures", "tutorial1_param_estimates_softmax.pdf"))
```

## Choice Rule Comparison

As introduced in the $M^3$ framework section, the choice rule determines how activations are converted to response probabilities. Because the choice rule affects both the parameter scale and the model's fit to data, selecting the appropriate rule is an important modeling decision. To compare the two options, we fit a second model with `choice_rule = "simple"` using the same formula and data. Under the simple rule, $a$ and $c$ are estimated on the log link, so the raw coefficients must be exponentiated to obtain activation values.

The posterior predictive checks for both models are qualitatively similar, but the softmax rule tends to capture the tails of the response distribution more accurately, particularly for the NPL category at larger set sizes. The simple rule slightly overpredicts other-list intrusions in some conditions. The companion script contains the full posterior predictive figures for both models.

For a formal comparison, we computed the marginal likelihood of each model via bridge sampling [@gronau2017bridge] and derive a Bayes factor. This requires `save_pars(all = TRUE)` during fitting (set in the `bmm()` call above):

```{r}
#| label: t1-bf-comparison
#| eval: false
bridge_softmax <- bridge_sampler(fit_m3_ss_softmax, repetitions = 10)
bridge_simple  <- bridge_sampler(fit_m3_ss_simple,  repetitions = 10)
bf(bridge_softmax, bridge_simple)
```

The Bayes factor strongly favored the softmax rule (median *BF*~10~ = `r bf10_fmt` across 10 bridge sampling repetitions). This is consistent with recent comparisons by @oberauer2023measurement and @li2026filtering, who found softmax to outperform the simple rule across different working memory tasks. We use the softmax rule as the default in the remaining tutorials.

## Hypothesis Testing

Because we fitted the model with `sample_prior = "yes"`, we can use `brms::hypothesis()` to test directed hypotheses and obtain evidence ratios (Savage-Dickey density ratios comparing posterior to prior density at the hypothesis boundary). For example, to test whether context binding exceeds item memory within the open-set experiment:

```{r}
#| label: t1-hypothesis
#| eval: false
hypothesis(fit_m3_ss_softmax, "c_expopenset = a_expopenset")
```

This returns a Savage-Dickey density ratio: the ratio of prior to posterior density at the point null $c = a$. Values substantially greater than 1 indicate that the data have shifted belief away from equality; values near 1 indicate inconclusive evidence. In the open-set experiment, the result is inconclusive (BF$_{10}$ = `r er_ca_fmt`), suggesting the data do not clearly distinguish whether $c$ and $a$ differ at the mean set size. We can similarly test whether set size reduces $c$ (averaged across experiments) and whether set size effects differ between experiments:

```{r}
#| label: t1-hypothesis-setsize
#| eval: false
# Comparion of c and a for the closed-set experiment
hypothesis(fit_m3_ss_softmax, "c_expclosedset = a_expclosedset")

# Set size effect on context binding (averaged across experiments)
hypothesis(fit_m3_ss_softmax,
  "(c_expopenset:ss_lin + c_expclosedset:ss_lin) / 2 = 0")

# Experiment difference in set size effect on c
hypothesis(fit_m3_ss_softmax,
  "c_expopenset:ss_lin = c_expclosedset:ss_lin")
```

The closed-set experiment yields strong evidence that $c > a$ (BF$_{10}$ = `r er_ca_closed_fmt`). However, there is moderate evidence is that set size reduces $a$ (BF$_{10}$ = `r er_ss_a_fmt`) and strong evidence for $c$ decreasing with larger set size (BF$_{10}$ = `r er_ss_c_fmt`). There is also strong evidence that the set size effect on $c$ is equal for the closed-set compared to the open-set experiment (BF$_{01}$ = `r er_exp_ss_c_fmt`). The companion script contains the full set of six hypothesis tests.

# Tutorial 2: Complex Span $M^3$

This tutorial extends the workflow to tasks where distractors appear alongside memoranda. We use data from @li2026filtering, who developed a complex-span filtering task to test whether processed information can be kept out of working memory. The companion script is `scripts/tutorial2_complex_span.R`; data preparation is in `scripts/prepare_Li2026_data.R`.

## What makes complex span different

The simple span model has three response categories. When distractors are present, and parameters capturing the effect of distractors should be estimated, two more categories are needed: the distractor paired with respect to the context of the tested target (*paired or close distractor*) and distractors from other positions (*other distractor*). The complex span (`cs`) model extends the activation formulas with a filtering parameter $f$ that determines how much activation distractors receive relative to memoranda. In the model for complex span tasks proposed by @oberauer2019simple it appears as a multiplier on item memory and context binding for distractor categories:

$$
\begin{aligned}
\text{A}_{\text{correct}} &= b + a + c \\
\text{A}_{\text{close dist.}} &= b + f \cdot a + f \cdot c \\
\text{A}_{\text{other}} &= b + a \\
\text{A}_{\text{other dist.}} &= b + f \cdot a \\
\text{A}_{\text{NPL}} &= b
\end{aligned}
$$

The parameter $f$ ranges from 0 to 1 (estimated on the logit scale). At the extremes, $f = 0$ means distractors receive only baseline activation, as if they were never encountered (perfect filtering); $f = 1$ means distractors are activated identically to memoranda, as if filtering failed entirely.

## Data

@li2026filtering presented participants with image pairs and manipulated whether targets and distractors were identified using an arrow cue pointing to one of the items before processing (pre-cue), after processing (retro-cue), or whether no distractors were present (control). We use Experiment 1 (45 participants after exclusions, within-subjects) for this tutorial. The data are pre-aggregated into response frequencies across the five categories:

```{r}
#| label: t2-data-load
#| eval: true
#| echo: false
data_agg <- read_csv(here("data", "Li_2026_ComplexSpan_Exp1_agg.csv"),
                     show_col_types = FALSE) %>%
  mutate(condition = factor(condition, levels = c("control", "pre", "retro")))

head(data_agg)
```

## Model specification

The model definition follows the same pattern as in Tutorial 1. As noted before, the `cs` version also assigns activation formulas by position: `resp_cats` must follow the order `corr, distc, other, disto, npl`. The model specification for the complex span version thus looks like this:

```{r}
#| label: t2-model-spec
m3_model_cs <- m3(
  resp_cats   = c("corr", "distc", "other", "disto", "npl"),
  num_options = c("n_corr", "n_distc", "n_other", "n_disto", "n_npl"),
  version     = "cs",
  choice_rule = "softmax"
)
```

Again, we can use the information from the model object that is returned by the `m3` call to obtain information on the parameters that are estimated and can be predicted. For the `cs` version, these are again context activation $c$, general activation $a$, and now also the filtering parameter $f$. According to the experimental design of experiment 1 from @li2026filtering, the predictor formula specifies condition effects on all three parameters with cell means coding and uncorrelated random effects per condition:

```{r}
#| label: t2-formula
m3_formula_cs <- bmf(
  c ~ 0 + condition + (0 + condition || participant),
  a ~ 0 + condition + (0 + condition || participant),
  f ~ 0 + condition + (0 + condition || participant)
)
```

There is one complication that we should address for this data: in the control condition, no distractors are presented (`n_distc = 0`, `n_disto = 0`), so $f$ does not influence the likelihood and is not identified. If left free, the sampler will wander through the prior, which can slow down sampling. The solution is to fix $f$ in the control condition using a `constant()` prior on both the fixed effect and the random effect SD:

```{r}
#| label: t2-priors
priors_cs <- c(
  prior(constant(0), nlpar = "f", coef = "conditioncontrol"),
  prior(constant(0),  class = "sd", nlpar = "f",
        coef = "conditioncontrol", group = "participant")
)
```

The specific values are not critical, as $f$ will not influence the likelihood in the control condition. Still, setting the fixed effect to 0 on the logit scale (effectively $f = 0.5$) and the SD to 0 pins the control-condition filtering at a fixed value. The remaining conditions still estimate $f$ freely. For more information on setting priors in `bmm`, you can check the `brms` documentation. As `bmm` builds on `brms`, the same syntax applies for setting priors. The companion script contains the full set of priors used for this model, which are mostly the default weakly informative priors.

## Model fitting and evaluation

Fitting follows the same `bmm()` call as in Tutorial 1, now additionally passing the custom priors. The model converged without issues ($\hat{R} \leq$ `r val_rhat_t2`, bulk ESS $>$ `r val_ess_t2`, no divergent transitions). @fig-t2-pp-check shows the posterior predictive check. The model captured the observed response proportions well across all three conditions and five categories.

```{r}
#| label: fig-t2-pp-check
#| eval: true
#| echo: false
#| fig-cap: "Posterior predictive check for the complex span $M^3$. Points show mean response proportions per category and condition; error bars show $\\pm 1$ SE across participants."
#| fig-alt: "Bar or point plot comparing observed and predicted response proportions for five categories (correct, paired distractor, other, other distractor, NPL) across control, pre-cue, and retro-cue conditions. Predicted values closely match observed proportions."
knitr::include_graphics(here("figures", "tutorial2_pp_check.pdf"))
```

## Parameter estimates and interpreting f

We now examine the posterior estimates for all three parameters. Item memory ($a$) and context binding ($c$) are on the identity link and can be read directly from the posterior draws. The filtering parameter $f$, however, is estimated on the logit scale. To interpret it as a proportional reduction between 0 and 1, we apply the inverse logit transformation (`plogis()`):

```{r}
#| label: t2-f-transform
#| eval: false
draws <- as_draws_df(fit_m3_cs) %>% as_tibble()
f_pre   <- plogis(draws$b_f_conditionpre)
f_retro <- plogis(draws$b_f_conditionretro)
```

```{r}
#| label: fig-t2-param-estimates
#| eval: true
#| echo: false
#| fig-cap: "Posterior parameter estimates for the complex span $M^3$. Left panels show item memory ($a$) and context binding ($c$) across conditions. Right panel shows distractor filtering ($f$) for pre-cue and retro-cue conditions, transformed to the probability scale. The control condition is omitted from the $f$ panel because it was fixed (no distractors present)."
#| fig-alt: "Multi-panel figure. Left panels show point-range plots of item memory (a) and context binding (c) across control, pre-cue, and retro-cue conditions. Right panel shows filtering parameter (f) on the probability scale for pre-cue and retro-cue, both around 0.65."
knitr::include_graphics(here("figures", "tutorial2_param_estimates.pdf"))
```

On the probability scale, $f$ is around 0.65 in both conditions (pre-cue: $M$ = `r val_f_pre_m`, 95% HDI [`r val_f_pre_lo`, `r val_f_pre_hi`]; retro-cue: $M$ = `r val_f_retro_m`, 95% HDI [`r val_f_retro_lo`, `r val_f_retro_hi`]), indicating some filtering. This matches the conclusion of @li2026filtering: once distractors are processed, they are encoded into working memory regardless of whether participants knew in advance which items were distractors.

## Hypothesis tests

Again, we can test condition differences for the different parameters formally using `brms::hypothesis()`:

```{r}
#| label: t2-hypothesis
#| eval: false
# Does f differ between pre-cue and retro-cue?
hypothesis(fit_m3_cs, "f_conditionpre - f_conditionretro = 0")

# Do a and c differ between control and distractor conditions?
hypothesis(fit_m3_cs,
  "a_conditioncontrol - (a_conditionpre + a_conditionretro) / 2 = 0")
```

The evidence ratio for the $f$ comparison indicated that filtering is equivalent between cue conditions (BF$_{01}$ = `r er_f_fmt`). There is inconclusive evidence for item memory $a$ comparing the distractor conditions to the control (BF$_{10}$ = `r er_a_dist_fmt`); however, context activation $c$ is lower in the distractor conditions (BF$_{10}$ = `r er_c_dist_fmt`).

## Comparison with the published model

The standard `cs` model makes a simplifying assumption: distractor attenuation is symmetric, with both item memory and context binding scaled by the single parameter $f$. This is a reasonable default, but it rules out the possibility that item and context information transfer to distractors at different rates. The original analysis by @li2026filtering tested exactly this distinction using a custom model with two separate ratio parameters, $r_a$ and $r_c$, allowing item and binding activation to transfer differently to distractors. Their findings suggest that item memory is barely reduced, whereas context activation is more affected. When a research question requires this kind of flexibility - or when the task design does not map onto any pre-built version - a custom model is needed. Tutorial 3 demonstrates how to define, validate, and fit such a model from scratch.


# Tutorial 3: Custom $M^3$ — Parameter Recovery

The previous tutorials used pre-built model versions (`ss` and `cs`) that required specific task designs and distractor categories to be present in the data. When a task design does not fit one of these templates, `bmm` allows defining a fully custom $M^3$ with user-specified response categories and activation formulas. This third tutorial demonstrates the custom version through a parameter recovery simulation: we define a custom model, simulate data with known parameters, fit the model, and check whether the generating values are recovered. The walkthrough follows the simplified script implementing a single simulation run `scripts/tutorial3_parameter_recovery_simple.R`; design-grid results from multiple repetitions for each design cell come from `scripts/tutorial3_parameter_recovery.R`.

## Why parameter recovery

Before applying a new or modified $M^3$ to empirical data, the model should be able to recover the parameters that generated the data. If recovery fails under ideal conditions (known generating values, no model misspecification), there are likely problems with the model specification or the design providing the data to identify the parameters. Recovery simulations are especially informative for custom models where the number of parameters and the functional form of the activation equations are chosen by the researcher [for a systematic treatment, see @goettmann2025recovery; for the broader rationale of simulation-based validation in Bayesian workflows, see @schad2021toward and @wilson2019ten].

## Defining a custom $M^3$

For this simulation we use a memory updating task as a running example to illustrate a common situation: the task produces response categories and activation structures that do not match either the `ss` or `cs` versions of the $M^3$. Specifically, replaced (outdated) items create additional response categories, and two experimental timing variables modulate activations in ways that require custom formulas. The task design mimicks the updating task and models from @oberauer2019simple. Participants memorize items at five positions, then position are updated randomly. At test, one position is probed and the participant selects from a recognition set containing: the current target (1 option), other current items (4), the old item from the tested position (1), old items from other positions (4), and not-presented lures (5). This produces five response categories, based on the model proposed by @oberauer2019simple for these kinds of updating tasks these categories have the following activation formulas:

$$
\begin{aligned}
\text{A}_{\text{correct}}  &= b + (1 + e \cdot t_e) \cdot (a + c) \\
\text{A}_{\text{other}}    &= b + (1 + e \cdot t_e) \cdot a \\
\text{A}_{\text{oldinpos}} &= b + \exp(-r \cdot t_r) \cdot d \cdot (1 + e \cdot t_e) \cdot (a + c) \\
\text{A}_{\text{otherold}} &= b + \exp(-r \cdot t_r) \cdot d \cdot (1 + e \cdot t_e) \cdot a \\
\text{A}_{\text{NPL}}      &= b
\end{aligned}
$$

Beyond the familiar $a$ (item memory) and $c$ (context binding) parameters, three new parameters appear: $d$ captures the residual activation of replaced (outdated) items that represents instant *deletion* of outdated items independent of the time manipulations, $e$ scales the benefit of extended encoding time $t_e$, and $r$ governs how old-item activation is removed with removal time $t_r$. To understand the composite terms: $(1 + e \cdot t_e)$ scales all memory activations as a function of encoding time - when encoding is brief the boost is minimal, when it is long activations are amplified. The term $\exp(-r \cdot t_r) \cdot d$ governs how much activation old items retain: $d$ sets the residual activation at the moment of replacement, and $r$ controls how quickly this residual is removed over time. Note that $t_e$ and $t_r$ are experimentally manipulated design variables (in ours simulation each at 0.25, 0.5, and 1.0 s), not estimated parameters, producing a $3 \times 3$ within-subjects design with 9 conditions.

As for fitting an $M^3$ in `bmm`, the model definition for simulatiing data for the $M^3$ specifies the response categories and option counts. In addition, we recommend to specify link functions to match each parameter's theoretical constraints: identity for $a$ and $c$ because activations are unbounded; logit for $d$ because it represents a proportion of original activation retained after replacement (bounded between 0 and 1); and log for $e$ and $r$ because encoding benefit and decay rate must be positive. These links functions are then internally applied to the parameters during sampleing:

```{r}
#| label: t3-model-spec
m3_model_custom <- m3(
  resp_cats = c("correct", "other", "oldinpos", "otherold", "npl"),
  num_options = c(
    "n_correct" = 1, "n_other" = 4, "n_oldinpos" = 1,
    "n_otherold" = 4, "n_npl" = 5
  ),
  version     = "custom",
  choice_rule = "softmax",
  links = list(
    a = "identity", c = "identity",
    d = "logit", e = "log", r = "log"
  )
)
```

As for model fitting, we have to specify the activation formulas separately using `bmf()`. For the simulation we do not need to add predictor formulas for the parameters as data generating parameters will be passed to the random generation function:

```{r}
#| label: t3-activation-formulas
m3_act_funs <- bmf(
  correct  ~ (1 + e * te) * (a + c) + b,
  other    ~ (1 + e * te) * a + b,
  oldinpos ~ exp(-r * tr) * d * (1 + e * te) * (a + c) + b,
  otherold ~ exp(-r * tr) * d * (1 + e * te) * a + b,
  npl      ~ b
)
```

Note that `te` and `tr` appear in the formulas but are not estimated parameters - they are data columns whose values vary across conditions. When `bmm` encounters variables in the activation formulas that are not listed as model parameters, it treats them as data columns and requires them to be present in the dataset. This allows custom models to incorporate any number of experimentally manipulated design variables directly into the activation equations, providing flexibility to capture complex task dynamics. For this simulation, we ensure that the simulated dataset includes `te` and `tr` columns with the appropriate values for each condition.

## Choosing generating parameter values

For the recovery simulation, we need group-level means and between-person standard deviations for each parameter to simulate subject-level parameters. These are specified on the link scale (the scale on which the sampler operates), not the native scale to respect the required bounds. The values were chosen to approximate empirical estimates from @li2025updating and to produce realistic individual differences. @tbl-gen-pars lists the generating values. The native range column shows the approximate span of individual differences ($\pm 2$ SD around the mean, transformed to native scale via the inverse link functions).

| Parameter | Link | Mean (link) | SD (link) | Native range ($\pm 2$ SD) |
|-----------|------|-------------|-----------|---------------------------|
| $c$ | identity | 2.0 | 0.50 | 1.0 -- 3.0 |
| $a$ | identity | 1.0 | 0.50 | 0.0 -- 2.0 |
| $d$ | logit | 0.85 | 0.65 | 0.38 -- 0.89 |
| $e$ | log | 0.00 | 0.35 | 0.50 -- 2.01 |
| $r$ | log | $-$0.69 | 0.38 | 0.24 -- 1.04 |

: Generating parameter values for the recovery simulation. All values are on the link scale. {#tbl-gen-pars}

Individual parameters are drawn from $\mathcal{N}(\mu_\text{link}, \sigma_\text{link})$ independently for each person and parameter. We simulated $N = 50$ participants with 25 observations per condition (225 observations total across the $3 \times 3$ design). Because the task uses set size 5 and we assumed that all positions within each trial are tested, each trial yields 5 responses, so 25 observations correspond to 5 trials per condition. With 9 conditions this amounts to 45 trials in total per participant, corresponding to roughly one hour of testing.

## Simulating data with rm3()

The `bmm` package provides the random generation function `rm3()` as the data-generating counterpart to the model-fitting workflow. Given a set of parameter values on the native scale, the model definition, and the activation formulas, it computes category probabilities via the specified choice rule and draws random samples from the multinomial distribution. Here is an example call for simulating a single participant in one condition:

```{r}
#| label: t3-rm3-demo
#| eval: false
demo_data <- rm3(
  n = 1, size = 25,
  pars = c(a = 1.2, c = 2.3, d = 0.7, e = 1.1, r = 0.5),
  m3_model = m3_model_custom,
  act_funs = m3_act_funs,
  te = 0.25, tr = 0.25
)
```

This returns a one-row data frame with response counts across the five categories. To simulate a full dataset, we apply `rm3()` over all 50 participants and 9 conditions (3 encoding times $\times$ 3 removal times), producing 450 rows of simulated data. The companion script available online contains the complete generation workflow and data inspection code.

## Model fitting

We fitted the simulated data with the same `bmm()` workflow already shown in Tutorial 1 and 2. For custom models, the prediction formula serves double duty: it repeats the activation equations (telling `bmm` how parameters combine into category activations) and appends regression formulas that specify how each parameter varies across participants (here, an intercept plus a random intercept per participant):

```{r}
#| label: t3-fit
#| eval: false
# Prediction formula: intercept + random intercept per participant
m3_pred_funs <- bmf(
  correct  ~ (1 + e * te) * (a + c) + b,
  other    ~ (1 + e * te) * a + b,
  oldinpos ~ exp(-r * tr) * d * (1 + e * te) * (a + c) + b,
  otherold ~ exp(-r * tr) * d * (1 + e * te) * a + b,
  npl      ~ b,
  a ~ 1 + (1 | participant), c ~ 1 + (1 | participant),
  d ~ 1 + (1 | participant), e ~ 1 + (1 | participant), r ~ 1 + (1 | participant)
)

fit <- bmm(
  formula = m3_pred_funs,
  model   = m3_model_custom,
  data    = sim_data,
  cores   = 4,
  backend = "cmdstanr",
  file    = here("output", "fit_m3_custom_simple_N50_T25")
)
```

## Recovery evaluation

Recovery can be evaluated at two levels. Group-level recovery asks whether the model recovers the true population means and between-person variability (i.e. standard deviation of the random intercepts and effects). Individual-level recovery asks whether each participant's true parameter value is correctly estimated - this is critical for researchers who plan to use $M^3$ parameter estimates as individual-difference measures, for example when correlating $a$ or $c$ with external variables.

### Group-level recovery

We first compare the recovered fixed effects to the true generating means. For each parameter, we extract the posterior median and 95% HDI from `fixef()` and check two things: (1) bias (is the posterior median close to the true value?) and (2) coverage (does the 95% HDI contain the true value?). We also extracted the between-person SDs from `brms::VarCorr()` and compare them to the generating SDs.

### Individual-level recovery

Individual-level recovery asks a similar question, but at the level of subject-level parameter estimates: for each participant, does the recovered estimate (fixed + random effect) match their true generating value? @fig-t3-indiv-scatter shows scatter plots of true versus recovered parameters for a single simulation cell ($N = 50$, 25 trials per condition). Points close to the diagonal indicate good recovery; the Pearson correlation ($r$) between true and estimated parameters is annotated in each panel.

```{r}
#| label: fig-t3-indiv-scatter
#| eval: true
#| echo: false
#| fig-cap: "Individual-level parameter recovery for the single simulation cell ($N = 50$, 25 trials per condition). Each point is one participant; the dashed line indicates perfect recovery. Pearson correlations are shown per parameter."
#| fig-alt: "Five scatter plots, one per parameter (a, c, d, e, r), showing true generating values on the x-axis and recovered posterior medians on the y-axis. Points for a, c, and d cluster near the diagonal; points for e and r show more scatter."
knitr::include_graphics(here("figures", "tutorial3_simple_indiv_scatter.pdf"))
```

Parameters $a$ and $c$ show the strongest individual-level recovery ($r$ = `r val_cor_a` and `r val_cor_c`, respectively). Parameters $d$ and $r$ show moderate recovery ($r$ = `r val_cor_d` and `r val_cor_r`). Recovery is weakest for $e$ ($r$ = `r val_cor_e`), which interacts multiplicatively with the design variable $t_e$, making it harder to separate from the other parameters. This is in line with more comprehensive parameter recovery simulations using the simple choice rule by @goettmann2025recovery, who found that $e$ and $r$ require considerably more data to recover well than $a$ and $c$.

## Practical guidance

The simplified simulation script demonstrates the workflow for a single design cell. To understand how recovery depends on sample size and the number of trials per condition, the full simulation script (`tutorial3_parameter_recovery.R`) varies both factors across a $3 \times 3$ grid: $N \in \{25, 50, 100\}$ and trials per condition $\in \{5, 10, 25\}$.

@fig-t3-group-recovery shows group-level recovery across the design grid. Recovered estimates are close to the true values in most cells, with HDIs narrowing as $N$ increases. With only 5 trials per condition, some parameters (particularly $e$ and $r$) show noticeable bias, but this improves substantially at 10 and 25 trials.

```{r}
#| label: fig-t3-group-recovery
#| eval: true
#| echo: false
#| fig-cap: "Group-level parameter recovery across the simulation design grid ($3 \\times 3$: sample sizes $\\times$ trials per condition). Points show recovered posterior medians; error bars show 95\\% HDIs. Dashed horizontal lines indicate the true generating value for each parameter. Gray point-ranges indicate estimates whose 95\\% HDI does not include the true value."
#| fig-alt: "Grid of point-range plots showing recovered group-level parameter estimates across 9 design cells (3 sample sizes by 3 trial counts per row facet). The x-axis shows sample size (25, 50, 100); dashed horizontal lines mark true values. HDIs narrow with larger N and more trials; some bias visible at 5 trials per condition. Gray estimates indicate non-coverage of the true value."
knitr::include_graphics(here("figures", "tutorial3_group_recovery.pdf"))
```

@fig-t3-heatmap shows individual-level recovery (Pearson correlation between true and recovered values) for each parameter across the same design grid. Individual-level recovery is more demanding than group-level recovery: $e$ and $r$ require both a reasonable sample size and a sufficient number of trials per condition to reach correlations above .7.

```{r}
#| label: fig-t3-heatmap
#| eval: true
#| echo: false
#| fig-cap: "Individual-level parameter recovery (Pearson $r$) across the simulation design grid. Each panel shows one parameter; rows are sample sizes and columns are trials per condition. Higher correlations (darker shading) indicate better recovery of individual differences."
#| fig-alt: "Heatmap with five panels (one per parameter). Cells show Pearson correlations between true and recovered individual parameters across 9 design cells. Parameters a and c show the highest correlations; d and r show moderate correlations; e shows the lowest correlations and requires more trials and larger samples."
knitr::include_graphics(here("figures", "tutorial3_recovery_heatmap.pdf"))
```

These results are based on a single replication per design cell. A proper simulation study should replicate each cell many times to account for Monte Carlo variability [see @goettmann2025recovery for a systematic approach, @chalmers2020simdesign for implementation tools, and @talts2018validating for a complementary calibration-based approach]. Recovery quality also depends on the model structure itself. Parameters that appear in interaction terms or that are only weakly identified by the design may require more data. As a rule of thumb, if individual-level correlations fall below .7 for a parameter of interest, the design likely needs more trials or a larger sample before that parameter can support individual-differences analyses. Running a recovery check before collecting empirical data is therefore a practical necessity for custom models: it reveals whether the planned sample size and trial count are sufficient, or whether the design needs refinement.

# Summary and Additional Resources

This tutorial covered three progressively complex applications of the $M^3$ framework using the `bmm` R package. The simple span example (Tutorial 1) introduced the full workflow: specification, fitting, evaluation, and hypothesis testing. It also compared the two choice rules implemented for the $M^3$. The complex span example (Tutorial 2) added distractor categories, the filtering parameter $f$, and the handling of non-identified parameters. The custom model example (Tutorial 3) showed how to define an $M^3$ from scratch for a memory updating task and how to validate it through parameter recovery.

A few practical recommendations. If the task produces three response categories (correct, other-list intrusion, NPL), the simple span version (`ss`) is the natural starting point. Tasks with distractors that are the same material type as memoranda can fit the complex span version (`cs`). For other designs, the custom version gives full control over activation formulas, parameters, and link functions. In all cases, the softmax choice rule is the recommended default based on current evidence. When defining a custom model, running a parameter recovery simulation before data collection reveals whether the planned design can distinguish the parameters of interest and whether the model is identified.

We hope that the implementation of the $M^3$ framework in the `bmm` package and the accompanying tutorials will make this powerful modeling approach accessible to a wide range of researchers interested in working memory. As noted before the $M^3$ is not necessarily limited to memory tasks, but can be applied to any decision paradigm yielding a multinomial response distribution [@mcfadden1974conditional; @busemeyer1993decision; @roe2001multialternative]. By providing a flexible yet structured way to model response distributions, $M^3$ can yield insights into the underlying cognitive processes that generate observed behavior. We encourage users to explore the package documentation, experiment with different model specifications, and contribute to the ongoing development of this tool.

## Additional resources

-   `bmm` package documentation and $M^3$ vignette: <https://venpopov.github.io/bmm/articles/bmm_m3.html>
-   Full companion scripts for all three tutorials: [repository link]
-   $M^3$ framework paper: @oberauer2019simple
-   Systematic parameter recovery for $M^3$: @goettmann2025recovery
-   Companion `bmm` tutorial for continuous-response mixture models: @frischkorn2025tutorial

<!-- ====================================================================== -->

# References {.unnumbered}

<!-- ====================================================================== -->

::: {#refs}
:::
