---
title: "Fitting the Memory Measurement Model (M3) with the bmm R Package: A Tutorial"
author:
  - name: Gidon T. Frischkorn
    orcid: 0000-0002-9555-0261
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Isabel Courage
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Chenyu Li
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
      - name: University of Chicago
        department: Department of Psychology
date: today
format:
  pdf:
    documentclass: apa7
    classoption: [man, 12pt]
    papersize: letter
    margin-left: 1in
    margin-right: 1in
    margin-top: 1in
    margin-bottom: 1in
    number-sections: false
    colorlinks: true
    fig-pos: "H"
    keep-tex: true
  docx:
    reference-doc: default
bibliography: references.bib
csl: apa.csl
abstract: |
  TODO: Draft abstract covering (a) the M3 framework and what it measures,
  (b) the bmm package as a new accessible implementation, (c) the three
  tutorials (simple span, complex span, custom model with parameter recovery),
  and (d) what readers will be able to do after completing the tutorials.
keywords:
  - working memory
  - measurement model
  - multinomial model
  - Bayesian estimation
  - R package
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here, bmm, brms, tidyverse, tidybayes, patchwork)
source(here("functions", "clean_plot.R"))
```

<!-- ====================================================================== -->
# Introduction
<!-- ====================================================================== -->

<!-- < 500 words per AMPPS guidelines. No extensive literature review. -->

<!-- Key points:
- WM researchers use categorical response tasks yielding response frequencies
- M3 framework (Oberauer & Lewandowsky, 2019) separates item memory (a) from
  context-specific binding (c) via competitive retrieval
- Previously required custom JAGS code — error-prone, barrier to adoption
- bmm R package provides standardized implementation: specify, fit, evaluate
  with minimal code; flexibly adapt to new designs
- This tutorial: three worked examples covering ss, cs, and custom versions
-->


<!-- ====================================================================== -->
# The M3 Model Framework
<!-- ====================================================================== -->

<!-- ~1,000–1,500 words. Teach the model, not the theory. -->

## What M3 Models

<!-- Categorical response data from memory tasks — response frequencies across
categories (correct, other-list intrusions, NPL, etc.) -->

## Activation Sources and Competitive Retrieval

<!-- Core mechanism: activation sources feed into competitive retrieval via a
choice rule (Luce's axiom). Response probability for a category depends on its
total activation relative to all categories.

Activation parameters:
  b = baseline (fixed for scaling)
  a = general item memory
  c = context-specific binding
  Additional parameters for custom models
-->

## Choice Rules

<!-- Simple (linear normalization) vs softmax (exponential normalization,
equivalent to n-AFC SDT). Brief explanation of when to prefer each; note that
recent evidence favors softmax. -->

## Model Versions

<!-- Three versions: ss (simple span, 3 categories), cs (complex span,
5 categories), custom (user-defined) -->

## The bmm Implementation

<!-- Brief overview: m3() creates the model, bmf() specifies predictor formulas,
bmm() fits via brms/Stan. Contrast with the old JAGS workflow.

Brief mention of relationship to MPT models (1–2 sentences). -->


<!-- ====================================================================== -->
# Tutorial 1: Simple Span M3
<!-- ====================================================================== -->

<!-- ~2,000–2,500 words. Full worked example with Oberauer (2019) data.
Companion script: scripts/tutorial1_simple_span.R -->

## Data

<!-- Oberauer (2019), Experiment 1 & 2. Simple span task with 3 response
categories: correct, other (intrusion from a different list position), NPL.
Data available at: https://osf.io/vekpd/ -->

```{r}
#| label: t1-data-load
data_agg <- read_csv(here("data", "Oberauer_2019_SimpleSpan_agg.csv"),
                     show_col_types = FALSE)
```

## Model Specification

<!-- m3(version = "ss") with softmax choice rule.
bmf() with condition effects on a and c. -->

```{r}
#| label: t1-model-spec
m3_model_ss <- m3(
  resp_cats   = c("corr", "other", "npl"),
  num_options = c("n_corr", "n_other", "n_npl"),
  version     = "ss",
  choice_rule = "softmax"
)
```

## Model Fitting

```{r}
#| label: t1-fit
#| eval: false
```

## Convergence and Model Evaluation

<!-- Rhat, ESS, posterior predictive checks -->

```{r}
#| label: fig-t1-pp-check
#| fig-cap: "Posterior predictive check for the simple span M3 model (softmax choice rule). Points show observed response proportions; error bars show ±1 SE."
# TODO: include pre-generated figure or generate inline
```

## Parameter Estimates

```{r}
#| label: fig-t1-param-estimates
#| fig-cap: "Posterior parameter estimates for the simple span M3 model (softmax choice rule)."
# TODO: include pre-generated figure or generate inline
```

## Choice Rule Comparison

<!-- Fit with both simple and softmax, compare results. Explain the difference
and why softmax is recommended as default. -->

## Hypothesis Testing

<!-- brms::hypothesis() with sample_prior = "yes" for Bayes Factors. -->


<!-- ====================================================================== -->
# Tutorial 2: Complex Span M3
<!-- ====================================================================== -->

<!-- ~1,500–2,000 words. Builds on Tutorial 1 workflow.
Data: Li, Frischkorn, & Oberauer (accepted, JEP:LMC).
Companion script: scripts/tutorial2_complex_span.R
Data preparation: scripts/prepare_Li2026_data.R -->

## What Makes Complex Span Different

<!-- Distractor categories, the f parameter.
5 response categories: corr, distc, other, disto, npl.
IMPORTANT: cs version assigns activation formulas by position — resp_cats
must follow the order corr, distc, other, disto, npl. -->

## Data

```{r}
#| label: t2-data-load
data_agg <- read_csv(here("data", "Li_2026_ComplexSpan_Exp1_agg.csv"),
                     show_col_types = FALSE)
```

## Model Specification

<!-- m3(version = "cs") with softmax choice rule.
Handling non-identified f in control condition (constant prior). -->

```{r}
#| label: t2-model-spec
m3_model_cs <- m3(
  resp_cats   = c("corr", "distc", "other", "disto", "npl"),
  num_options = c("n_corr", "n_distc", "n_other", "n_disto", "n_npl"),
  version     = "cs",
  choice_rule = "softmax"
)
```

## Model Fitting and Evaluation

```{r}
#| label: fig-t2-pp-check
#| fig-cap: "Posterior predictive check for the complex span M3 model."
# TODO: include pre-generated figure
```

## Parameter Estimates

```{r}
#| label: fig-t2-param-estimates
#| fig-cap: "Posterior parameter estimates for the complex span M3 model. Left panels show item memory (a) and context binding (c) across conditions. Right panel shows distractor filtering (f) for pre-cue and retro-cue conditions."
# TODO: include pre-generated figure
```

## Interpreting the f Parameter

<!-- f = 0 → perfect filtering; f = 1 → no filtering.
Compare pre-cue vs retro-cue. -->

## Hypothesis Tests

<!-- brms::hypothesis() for condition comparisons. -->

## Comparison with the Published Model

<!-- Published analysis used separate ra/rc parameters (different attenuation
for item vs binding activation). The cs model constrains ra = rc = f.
When this constraint is too restrictive → custom model needed (Tutorial 3). -->


<!-- ====================================================================== -->
# Tutorial 3: Custom M3 — Parameter Recovery
<!-- ====================================================================== -->

<!-- ~2,500–3,000 words. Centerpiece for flexibility.
Inline code follows: scripts/tutorial3_parameter_recovery_simple.R
Design grid results from: scripts/tutorial3_parameter_recovery.R -->

## Why Custom Models

<!-- When standard ss/cs don't fit the experimental design. The custom version
allows specifying M3 models of arbitrary complexity with user-defined
activation formulas, as long as the model is identified. -->

## Defining a Custom M3 Model

<!-- Memory updating model (Oberauer & Lewandowsky, 2019) with 5 response
categories and 5 estimated parameters.

Activation formulas:
  correct  = b + (1 + e*te) * (a + c)
  other    = b + (1 + e*te) * a
  oldinpos = b + exp(-r*tr) * d * (1 + e*te) * (a + c)
  otherold = b + exp(-r*tr) * d * (1 + e*te) * a
  npl      = b
-->

```{r}
#| label: t3-model-spec
m3_model_custom <- m3(
  resp_cats = c("correct", "other", "oldinpos", "otherold", "npl"),
  num_options = c(
    "n_correct" = 1, "n_other" = 4, "n_oldinpos" = 1,
    "n_otherold" = 4, "n_npl" = 5
  ),
  version     = "custom",
  choice_rule = "softmax",
  links = list(
    a = "identity", c = "identity",
    d = "logit", e = "log", r = "log"
  )
)
```

```{r}
#| label: t3-activation-formulas
m3_act_funs <- bmf(
  correct  ~ (1 + e * te) * (a + c) + b,
  other    ~ (1 + e * te) * a + b,
  oldinpos ~ exp(-r * tr) * d * (1 + e * te) * (a + c) + b,
  otherold ~ exp(-r * tr) * d * (1 + e * te) * a + b,
  npl      ~ b
)
```

## Choosing Generating Parameter Values

<!-- Group-level means and between-person SDs on the link scale.
Explain link-to-native scale conversion. -->

## Simulating Data with rm3()

<!-- Demonstrate rm3() for a single participant and condition first,
then show the full simulation loop. -->

```{r}
#| label: t3-rm3-demo
# Simulate for one participant, one condition
demo_data <- rm3(
  n = 1, size = 50, pars = pars_p1,
  m3_model = m3_model_custom, act_funs = m3_act_funs,
  te = 0.25, tr = 0.25
)
```

## Model Fitting

<!-- Fit the same model to simulated data. -->

## Recovery Evaluation

### Group-Level Recovery

```{r}
#| label: fig-t3-group-recovery
#| fig-cap: "Group-level parameter recovery across the simulation design grid. Points show recovered estimates; error bars show 95% credible intervals. Dashed line indicates perfect recovery."
# TODO: include pre-generated figure (tutorial3_group_recovery.pdf)
```

### Individual-Level Recovery

```{r}
#| label: fig-t3-indiv-scatter
#| fig-cap: "Individual-level parameter recovery for the single simulation cell (N = 50, 50 trials per condition). Each point represents one participant. Dashed line indicates perfect recovery."
# TODO: include pre-generated figure (tutorial3_simple_indiv_scatter.pdf)
```

## Practical Guidance

<!-- How sample size and trials per condition affect recovery quality.
Note on Monte Carlo noise: single replication demonstrates the workflow;
proper simulation studies should replicate each cell.
Reference Göttmann et al. and SimDesign. -->


<!-- ====================================================================== -->
# Summary and Additional Resources
<!-- ====================================================================== -->

<!-- ~300–500 words. Per AMPPS format — brief summary, no traditional Discussion. -->

<!-- Recap: three model versions, basic workflow, choice rules, custom flexibility.
Practical recommendations: which version to choose, when custom, importance
of parameter recovery. -->

## Additional Resources

<!-- - bmm documentation: https://venpopov.github.io/bmm/articles/bmm_m3.html
     - Full tutorial scripts: [repository link]
     - Parameter recovery methodology: Göttmann et al. (n.d.)
     - M3 framework paper: Oberauer & Lewandowsky (2019)
     - bmm mixture model tutorial: Popov et al. (2025) -->


<!-- ====================================================================== -->
# References {.unnumbered}
<!-- ====================================================================== -->

::: {#refs}
:::
