---
title: "Fitting the Memory Measurement Model ($M^3$) with the bmm R Package: A Tutorial"
shorttitle: "Fitting the $M^3$ with the bmm R Package"
author:
  - name: Gidon T. Frischkorn
    orcid: 0000-0002-5055-9764
    email: gidon.frischkorn@psychologie.uzh.ch
    corresponding: true
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Isabel Courage
    orcid: 0009-0000-8493-2240
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Chenyu Li
    orcid: 0000-0001-8815-6189
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
      - name: University of Chicago
        department: Department of Psychology
author-note:
  disclosures:
    study-registration: "This study was not preregistered."
    data-sharing: "Data, analysis code, and companion R scripts for all three tutorials are available at [repository link]."
    conflict-of-interest: "The authors declare no conflicts of interest."
    financial-support: "[FUNDING INFORMATION]"
    gratitude: "[ACKNOWLEDGMENTS]"
    authorship-agreements: |
      Gidon T. Frischkorn: Conceptualization, Methodology, Software, Formal Analysis, Visualization, Writing -- Original Draft.
      Isabel Courage: Software, Formal Analysis, Writing -- Review & Editing.
      Chenyu Li: Software, Data Curation, Writing -- Review & Editing.
format:
  apaquarto-pdf:
    keep-tex: true
  apaquarto-docx: default
bibliography: references.bib
abstract: |
  The memory measurement model ($M^3$) decomposes categorical response data from memory tasks into latent activation parameters, separating item memory from context-specific binding via competitive retrieval. Fitting $M^3$ has so far required custom JAGS or Stan code for each task design. The bmm R package provides a standardized interface to specify, fit, and evaluate Bayesian hierarchical $M^3$ with minimal code. This tutorial demonstrates the bmm workflow through three worked examples: fitting a simple span $M^3$ to cued recall data, where Bayes factor model comparison favors the softmax over the simple choice rule; extending the workflow to a complex span task with distractor categories and a filtering parameter; and defining a fully custom $M^3$ for a memory updating task, validating it through parameter recovery across a range of sample sizes and trial counts. Complete R scripts accompany each tutorial. After completing the tutorials, readers will be able to fit standard and custom $M^3$ to their own data and validate new model specifications through simulation.
keywords:
  - working memory
  - measurement model
  - multinomial model
  - Bayesian estimation
  - R package
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
#| eval: true
pacman::p_load(here, bmm, brms, tidyverse, tidybayes, patchwork)
source(here("functions", "clean_plot.R"))

# ---- Load fitted models for inline reporting ----
fit_m3_ss_softmax <- readRDS(here("output", "fit_m3_ss_softmax.rds"))
fit_m3_ss_simple  <- readRDS(here("output", "fit_m3_ss_simple.rds"))
fit_m3_cs         <- readRDS(here("output", "fit_m3_cs.rds"))
bridge_softmax    <- readRDS(here("output", "bridge_m3_ss_softmax.rds"))
bridge_simple     <- readRDS(here("output", "bridge_m3_ss_simple.rds"))

# ---- Convergence diagnostics across all models ----
get_diagnostics <- function(fit) {
  s <- summary(fit)
  rhats <- s$fixed[, "Rhat"]
  ess   <- s$fixed[, "Bulk_ESS"]
  for (g in names(s$random)) {
    rhats <- c(rhats, s$random[[g]][, "Rhat"])
    ess   <- c(ess,   s$random[[g]][, "Bulk_ESS"])
  }
  list(rhat = max(rhats, na.rm = TRUE), ess = min(ess, na.rm = TRUE))
}

diag_sm <- get_diagnostics(fit_m3_ss_softmax)
diag_si <- get_diagnostics(fit_m3_ss_simple)
diag_cs <- get_diagnostics(fit_m3_cs)

val_rhat_all <- sprintf("%.3f", max(diag_sm$rhat, diag_si$rhat, diag_cs$rhat))
val_ess_all  <- format(round(min(diag_sm$ess, diag_si$ess, diag_cs$ess)),
                       big.mark = ",")
val_rhat_t2  <- sprintf("%.3f", diag_cs$rhat)
val_ess_t2   <- format(round(diag_cs$ess), big.mark = ",")

# ---- Bayes factor: softmax vs simple ----
bf_result   <- bridgesampling::bf(bridge_softmax, bridge_simple)
bf10_choice <- bf_result$bf

bf10_fmt <- if (min(bf10_choice) >= 10000) {
  ev <- floor(log10(bf10_choice))
  cv <- bf10_choice / 10^ev
  sprintf("$%.1f \\times 10^{%d}$", cv, ev)
} else {
  sprintf("%.1f", min(bf10_choice))
}

# ---- Evidence ratios ----
fmt_er <- function(x) {
  if (is.na(x)) return("NA")
  if (is.infinite(x)) return("> 10,000")
  if (x >= 1000) return(format(round(x), big.mark = ","))
  if (x >= 10) return(sprintf("%.1f", x))
  sprintf("%.2f", x)
}

h_ca      <- hypothesis(fit_m3_ss_softmax, "c_expopenset - a_expopenset > 0")
er_ca_fmt <- fmt_er(h_ca$hypothesis$Evid.Ratio)

h_f      <- hypothesis(fit_m3_cs, "f_conditionpre - f_conditionretro = 0")
er_f_fmt <- fmt_er(h_f$hypothesis$Evid.Ratio)

# ---- Tutorial 2: f on probability scale ----
draws_cs <- as_draws_df(fit_m3_cs) |> as_tibble()

f_pre_hdci <- tibble(f = plogis(draws_cs$b_f_conditionpre)) |>
  mean_hdci(f, .width = 0.95)
f_retro_hdci <- tibble(f = plogis(draws_cs$b_f_conditionretro)) |>
  mean_hdci(f, .width = 0.95)

val_f_pre_m    <- sprintf("%.2f", f_pre_hdci$f)
val_f_pre_lo   <- sprintf("%.2f", f_pre_hdci$.lower)
val_f_pre_hi   <- sprintf("%.2f", f_pre_hdci$.upper)
val_f_retro_m  <- sprintf("%.2f", f_retro_hdci$f)
val_f_retro_lo <- sprintf("%.2f", f_retro_hdci$.lower)
val_f_retro_hi <- sprintf("%.2f", f_retro_hdci$.upper)

# ---- Tutorial 3: Recovery correlations ----
# Reproduce true generating parameters using the same seed as the script
set.seed(2025)
N_sim <- 50
true_pars_t3 <- tibble(
  participant = 1:N_sim,
  c_link = rnorm(N_sim, 2.0, 0.50),
  a_link = rnorm(N_sim, 1.0, 0.50),
  d_link = rnorm(N_sim, 0.85, 0.65),
  e_link = rnorm(N_sim, 0.00, 0.35),
  r_link = rnorm(N_sim, -0.69, 0.38)
)

t3_file <- here("output", "fit_m3_custom_simple_N50_T25.rds")
if (file.exists(t3_file)) {
  fit_t3 <- readRDS(t3_file)
  co_t3  <- coef(fit_t3)
  par_names_t3 <- c("a", "c", "d", "e", "r")
  indiv_corrs <- sapply(par_names_t3, function(p) {
    est <- co_t3$participant[, , paste0(p, "_Intercept")]
    cor(true_pars_t3[[paste0(p, "_link")]], est[, "Estimate"])
  })
} else {
  indiv_corrs <- setNames(rep(NA_real_, 5), c("a", "c", "d", "e", "r"))
}

val_cor_a <- sprintf("%.2f", indiv_corrs["a"])
val_cor_c <- sprintf("%.2f", indiv_corrs["c"])
val_cor_d <- sprintf("%.2f", indiv_corrs["d"])
val_cor_e <- sprintf("%.2f", indiv_corrs["e"])
val_cor_r <- sprintf("%.2f", indiv_corrs["r"])
```

# Introduction

Observed performance in memory and decision tasks is often summarized by overall accuracy. Yet, for many tasks accuracy collapses across qualitatively distinct response categories and error types that may arise from different latent processes. In particular, different error types are diagnostically informative because different cognitive processes imply different conditional response probabilities over observable outcomes. Multinomial processing tree (MPT) models formalize this logic by treating categorical responses—including distinct error classes—as consequences of latent states such as successful retrieval, guessing, or source confusion [@batchelder1999theoretical; @erdfelder2009multinomial]. Similarly, process-dissociation approaches demonstrate that patterns of correct and incorrect responses across task constraints can separate the contributions of multiple processes within a single task [@jacoby1991process; @yonelinas2012process]. In sum, the full multinomial response distribution - not accuracy alone - carries more information about latent cognitive mechanisms than a single accuracy score.

Working memory research frequently relies on tasks that inherently yield response frequencies across discrete outcome categories, such as correct responses, feature confusions, location errors, or intrusions [@oberauer2018benchmarks]. For discrete feature spaces (e.g., words, numbers, letters), these data are naturally multinomial: each retrieval produces one of several mutually exclusive response types. Thus, theoretical claims can often be evaluated by how experimental manipulations and individual differences redistribute probability mass across these categories.

The recently proposed memory measurement model [$M^3$; @oberauer2019simple] framework provides one such decomposition. Its core assumptions are that 1) multiple activation sources contribute to each response category and that  2) selection among categories follows a competitive retrieval rule [@luce1959individual]. Although the $M^3$ framework was introduced for working memory tasks, the general architecture is domain-general: whenever a task yields categorical responses that can be traced to distinct latent activation sources driving competitive selection, the framework applies.[^1]

[^1]: Readers familiar with multinomial processing tree (MPT) models will note similarities: both frameworks decompose categorical response distributions into latent process parameters. However, whereas MPT models use discrete processing trees with binary branching probabilities [@batchelder1999theoretical; e.g., @heck2018treebugs for Bayesian hierarchical implementations], $M^3$ uses continuous activation values with competitive selection, making it particularly suited for theories emphasizing graded memory strength [for a detailed comparison, see @oberauer2019simple].

Because the $M^3$ framework is flexible — @oberauer2019simple already introduced more than five model architectures — the activation formulas that drive response selection typically vary across tasks and response categories. Each new application therefore required translating its model into custom JAGS or Stan code, making implementations error-prone and hard to reuse. The `bmm` R package [@bmm2025] addresses this limitation by providing a standardized workflow to specify, fit, and evaluate Bayesian hierarchical $M^3$ with minimal code, while remaining flexible enough to accommodate new task designs and custom model specifications. In this tutorial, we illustrate this approach with three worked examples covering the standard simple span (ss) and complex span (cs) tasks, and a custom extension.

# The $M^3$ Framework

This section briefly introduces the core concepts of the $M^3$ framework — activation formulas, choice rules, and model versions — that apply across all implementations.

Consider a cued recall task (e.g., remembering letters in serial positions or icons in spatial positions) in which a participant sees a retrieval cue and selects a response from a set containing the correct target, other studied items, and new lures. Each response falls into one of three categories: *correct* — the item associated with the retrieval cue, *other-list intrusion* — a studied item associated with a different cue, or *not-presented lure* (NPL) — an item not presented during the trial. The $M^3$ framework explains each category's selection frequency by specifying activation formulas: different activation sources feed into each category, and response probabilities follow from competitive selection among all options.

## Activation Formulas and Model Parameters

Each response category's activation formula specifies which latent sources contribute to its selection probability. Because the $M^3$ framework is not committed to any particular memory theory, different assumptions about the processes limiting memory performance can be accommodated through different activation formulas.

In the $M^3$ framework, each response category receives activation from one or more latent sources. For the cued recall task introduced above, the theoretically assumed activation sources are:

-   $b$: **baseline activation**, the tendency to select any response option in the absence of memory-based evidence. This parameter is fixed (not estimated; e.g., $b = 0$ for the softmax rule) to set the scale of the model.
-   $a$: **item activation**, or general memory strength for studied items. An item that was encountered during study receives this activation regardless of its position or context.
-   $c$: **context binding**, the strength of the association between an item and its specific study context (e.g., serial position, paired associate).

From these parameters, we can specify how they are combined to produce total activations for each response category. The simple span model proposed by @oberauer2019simple assumes that these activation sources combine additively, with the correct target receiving activation from all three sources, other studied items receiving item activation but lacking context binding, and lures receiving only baseline activation:

$$
\begin{aligned}
\text{A}_{\text{correct}} &= b + a + c \\
\text{A}_{\text{other}} &= b + a \\
\text{A}_{\text{lures}} &= b
\end{aligned}
$$

Another perspective on the interplay of these processes could also assume that context activation boosts item activation, such that the correct target receives a multiplicative boost from context binding, while other studied items receive only item activation:

$$
\begin{aligned}
\text{A}_{\text{correct}} &= b + a \cdot (1 + c) \\
\text{A}_{\text{other}} &= b + a \\
\text{A}_{\text{lures}} &= b
\end{aligned}
$$

These two examples illustrate how distinct theoretical perspectives can be encoded through the activation formulas. The `bmm` package provides pre-specified formulas for the simple span and complex span versions proposed by @oberauer2019simple, and also allows users to define custom formulas for novel task designs.

## Competitive Selection and Choice Rules

To link activations to observed responses, each category's activation must be converted into a response probability. The $M^3$ framework assumes competitive selection: a category's probability depends on its activation relative to all competitors. Because the choice rule requires knowing the number of options per category, tasks designed for $M^3$ should use a recognition format in which the experimenter controls the available response options.

Given these option counts, $M^3$ provides two choice rules that differ in how activations are transformed before normalization. The *simple* (linear) choice rule [@luce1959individual] normalizes activations directly:

$$P(k) = \frac{n_k \cdot \text{A}_k}{\sum_{j=1}^{K} n_j \cdot \text{A}_j}$$

where $\text{A}_k$ is the total activation for category $k$ and $n_k$ is the number of response options in that category. This rule requires all activations to be positive, so baseline $b$ is fixed at a small positive value (0.1 by default).

The *softmax* (exponential) choice rule applies an exponential transformation before normalization:

$$P(k) = \frac{n_k \cdot \exp(\text{A}_k)}{\sum_{j=1}^{K} n_j \cdot \exp(\text{A}_j)}$$

Because $\exp(x) > 0$ for all $x$, softmax guarantees valid probabilities regardless of activation values. The baseline is fixed at $b = 0$ (since $\exp(0) = 1$, lures receive a positive baseline contribution automatically).

The softmax rule is formally equivalent to an $n$-alternative forced-choice signal detection model with Gumbel-distributed noise [@thurstone1927law; @yellott1977relationship]. Under this interpretation, each category's activation represents a deterministic signal, independent Gumbel noise is added to each alternative, and the observer selects the category with the highest noisy value. This equivalence gives activation parameters a principled interpretation as discriminability measures. Recent empirical comparisons have favored the softmax rule over the simple rule for working memory data [@oberauer2023measurement; @li2026filtering] although the simple choice rule was endorsed initially [@oberauer2019simple]. Tutorial 1 compares both rules; Tutorials 2 and 3 use the softmax rule as the default.

## Model Versions

The `bmm` package provides three $M^3$ versions, two of which implement canonical paradigms often used in working memory research. The versions differ in the response categories and estimated parameters. Here we shortly introduce them, the following tutorials provide more details.

The **simple span** (`ss`) version implements $M^3$ for cued recall tasks separating three response categories (correct response, other, lures) with two free parameters ($a$, $c$). It implements the additive activation formula proposed by @oberauer2019simple.

The **complex span** (`cs`) version extends the simple span to tasks that interleave distractor processing with memorization. It requires that memoranda and distractors are drawn from the same material so that distractors can plausibly intrude during retrieval [for examples, see @oberauer2019simple; @li2026filtering]. The `cs` version separates five response categories (correct, paired distractor, other memorandum, other distractor, and NPL) with three free parameters ($a$, $c$, $f$). The additional parameter $f$ captures distractor filtering efficiency on a scale from 0 to 1: $f = 0$ indicates perfect filtering (distractors receive only baseline activation), while $f = 1$ indicates no filtering (distractors are activated as strongly as memoranda).

The **custom** version allows user-defined response categories, activation formulas, and parameters. This version allows researchers to specify $M^3$ of arbitrary complexity for novel task designs, as long as the model remains identified. The `bmm` package provides a flexible syntax for defining custom models, and Tutorial 3 demonstrates how to validate a custom model through parameter recovery simulations.

## The bmm Workflow

Fitting $M^3$ with the `bmm` R package [@bmm2025] follows a three-step workflow that is consistent for all measurement models implemented in the package:

1.  **Define the model**: The package includes a constructor function `m3()` that generates a model object with all relevant information for model specification. This entails the variable names of response categories in the data, the number of response options per category, the model version, and the choice rule.
2.  **Specify predictor formulas**: The `bmm` package provides a `bmmformula()` or shorthand `bmf()` function to specify the linear prediction formulas that define how experimental conditions and individual differences predict each activation parameter. The `bmmformula` syntax uses the same formula syntax as `brms` [@burkner2017brms] and also includes random effects and non-linear formulas.[^2]
3.  **Fit the model**: The model object and prediction formula are passed to `bmm()` together with the data. This function translates the specification into a `brms` model and passes it to Stan [@carpenter2017stan] for Bayesian estimation.

[^2]: For details on the `bmmformula` syntax, see <https://venpopov.com/bmm/articles/bmm_bmmformula.html>.

Because `bmm` builds on `brms`, fitted models inherit most of the `brms` post-processing ecosystem, including posterior predictive checks, hypothesis tests, and model comparison.[^3]

[^3]: We are continously working on extending the post-processing functions to accommodaate the specific requirements of the measurement models implemented in the `bmm` R package. 

# Tutorial 1: Simple Span $M^3$

This tutorial walks through the complete $M^3$ workflow using data from @oberauer2019bindings, who measured cued recall with word pairs. Two experiments (20 participants each) differed in stimulus set type: Experiment 1 used new word pairs on each trial (open set), while Experiment 2 reused a small set of words across trials (closed set). Set size (2, 4, 6, 8 pairs) varied within participants. The full companion script (`scripts/tutorial1_simple_span.R`) contains all code; we reproduce the key steps here.

## Data

Each recognition trial presents a probe together with a set of response options: the correct target, other studied items from the same list, and not-presented lures. Participants select one item, producing a categorical response: *correct*, *other* (intrusion from a different list position), or *NPL* (not-presented lure). The number of response options per category varies across the 24 recognition conditions (combinations of set size, list items in the recognition set, and number of lures).

$M^3$ in `bmm` expects aggregated data: response frequency counts per participant and condition, along with the number of response options per category. Starting from the trial-level data (available at https://osf.io/vekpd/), we filter to recognition trials, aggregate response categories, and compute option counts:

```{r}
#| label: t1-data-prep
# Read both experiments; create unique participant IDs
col_names <- c("id", "session", "block", "trial", "setsize",
               "rsizeList", "rsizeNPL", "tested", "response", "rcat", "rt")

read_exp <- function(file, exp_label) {
  read.table(here("data", file), header = FALSE,
             col.names = col_names, fill = TRUE) %>%
    drop_na() %>%
    mutate(exp = exp_label,
           id  = paste(id, exp_label, sep = "_"))
}

data_raw <- bind_rows(
  read_exp("Oberauer_2019_SimpleSpan_Exp1.dat", "openset"),
  read_exp("Oberauer_2019_SimpleSpan_Exp2.dat", "closedset")
)
```

```{r}
#| label: t1-data-agg
# Keep recognition trials only, aggregate to response frequencies
data_agg <- data_raw %>%
  filter(rsizeList + rsizeNPL > 0) %>%
  group_by(id, exp, setsize, rsizeList, rsizeNPL) %>%
  summarise(corr  = sum(rcat == 1),
            other = sum(rcat == 2),
            npl   = sum(rcat == 3), .groups = "drop") %>%
  mutate(n_corr  = 1,
         n_other = rsizeList - 1,  # rsizeList includes the correct item
         n_npl   = rsizeNPL,
         ss_lin  = setsize - mean(setsize),
         exp     = factor(exp))
```

The resulting dataset has 960 rows (40 participants $\times$ 24 recognition conditions) with response frequencies (`corr`, `other`, `npl`) and the corresponding option counts (`n_corr`, `n_other`, `n_npl`). Set size is mean-centered (`ss_lin`) for the linear predictor.

## Model Specification

Specifying an $M^3$ requires two objects: a model definition (`m3()`) and a predictor formula (`bmf()`). The model definition declares the response categories, the columns containing option counts, the model version, and the choice rule. For the simple span version with the softmax rule:

```{r}
#| label: t1-model-spec
m3_model_ss <- m3(
  resp_cats   = c("corr", "other", "npl"),
  num_options = c("n_corr", "n_other", "n_npl"),
  version     = "ss",
  choice_rule = "softmax"
)
```

The predictor formula specifies how experimental conditions map onto the activation parameters $a$ and $c$. Because the two experiments are between-subjects, we use cell means coding (`0 + exp`) so that each experiment receives its own intercept rather than a difference from a reference level. Set size enters as a linear predictor (`exp:ss_lin`), and random effects allow both intercepts and slopes to vary across participants, with separate variance components per experiment (`gr(id, by = exp)`):

```{r}
#| label: t1-formula
m3_formula <- bmf(
  c ~ 0 + exp + exp:ss_lin + (1 + ss_lin | gr(id, by = exp)),
  a ~ 0 + exp + exp:ss_lin + (1 + ss_lin | gr(id, by = exp))
)
```

Under the softmax rule, $a$ and $c$ are estimated on the identity link (native activation scale). The default priors can be inspected with `default_prior(m3_formula, m3_model_ss, data = data_agg)`. We use the defaults throughout this tutorial; Tutorial 2 shows how to set custom priors when needed.

## Model Fitting

The `bmm()` function combines the model definition and formula, passes them to `brms` and Stan, and returns a fitted model object. All models in this tutorial were estimated with 4 chains of 2,000 iterations each (1,000 warmup), using `cmdstanr` as the backend. We set `sample_prior = "yes"` so that prior samples are available for later hypothesis tests. The `file` argument caches the fitted model to disk; re-running the script loads the cached fit rather than re-fitting.

```{r}
#| label: t1-fit
#| eval: false
fit_m3_ss_softmax <- bmm(
  formula      = m3_formula,
  model        = m3_model_ss,
  data         = data_agg,
  cores        = 4,
  backend      = "cmdstanr",
  sample_prior = "yes",
  save_pars    = save_pars(all = TRUE),
  file         = here("output", "fit_m3_ss_softmax")
)
```

After fitting, check convergence with `summary(fit_m3_ss_softmax)`. All $\hat{R}$ values [@gelman1992inference] should be below 1.01 and effective sample sizes should be adequate. For all models in this tutorial, these criteria were met ($\hat{R} \leq$ `r val_rhat_all`, bulk ESS $>$ `r val_ess_all`). For guidance on Bayesian convergence diagnostics, see @gabry2019visualization and the `brms` documentation.

## Model Evaluation

A posterior predictive check compares observed response proportions to those predicted by the fitted model. We use `posterior_epred()` to obtain expected counts per category for each observation, average across posterior draws, and aggregate to participant-level proportions by set size and experiment. @fig-t1-pp-check shows the result. Across all conditions and response categories, predicted proportions fall within one standard error of the observed means.

```{r}
#| label: fig-t1-pp-check
#| fig-cap: "Posterior predictive check for the simple span $M^3$ (softmax choice rule). Points and lines show mean response proportions by set size and experiment; error bars show $\\pm 1$ SE across participants. Color distinguishes observed from predicted values."
#| fig-alt: "Line plot comparing observed and predicted response proportions across set sizes 2, 4, 6, and 8 for three response categories (correct, other, NPL) in open-set and closed-set experiments. Predicted proportions closely track observed values in all conditions."
knitr::include_graphics(here("figures", "tutorial1_pp_check_softmax.pdf"))
```

## Parameter Estimates

With the softmax choice rule, $a$ and $c$ are on the identity link, so the posterior draws are directly interpretable as activation values. To compute condition-specific estimates, we reconstruct the linear predictor from the fixed effects: for each set size $s$, $\hat{a}_{\text{exp}} = \beta_{a,\text{exp}} + \beta_{a,\text{exp:ss\_lin}} \cdot (s - \bar{s})$, and analogously for $c$. The companion script extracts draws with `as_draws_df()` and computes 95% highest-density intervals [HDIs; @mcelreath2020statistical] via `tidybayes::mean_hdci()`.

@fig-t1-param-estimates shows the resulting estimates. Context binding ($c$) is consistently higher than item memory ($a$), which suggests that the primary source of correct responding is context-specific retrieval rather than general familiarity. Both parameters decrease with set size, but the decline is steeper for $c$, consistent with binding being more sensitive to memory load. The two experiments show similar patterns, though the closed set tends to produce slightly higher item activation, likely because the same words recur across trials and accumulate familiarity.

```{r}
#| label: fig-t1-param-estimates
#| fig-cap: "Posterior estimates of item memory ($a$) and context binding ($c$) across set sizes for the open-set and closed-set experiments (softmax choice rule). Points show posterior means; error bars show 95\\% HDIs."
#| fig-alt: "Point-range plot showing posterior means and 95% HDIs for item memory (a) and context binding (c) across set sizes 2, 4, 6, and 8, separately for open-set and closed-set experiments. Both parameters decrease with set size, with c consistently higher than a."
knitr::include_graphics(here("figures", "tutorial1_param_estimates_softmax.pdf"))
```

## Choice Rule Comparison

To compare the two choice rules, we fit a second model with `choice_rule = "simple"` using the same formula and data. Under the simple rule, $a$ and $c$ are estimated on the log link, so the raw coefficients must be exponentiated to obtain activation values.

The posterior predictive checks for both models are qualitatively similar, but the softmax rule tends to capture the tails of the response distribution more accurately, particularly for the NPL category at larger set sizes. The simple rule slightly overpredicts other-list intrusions in some conditions. These figures are available in the companion script.

For a formal comparison, we compute the marginal likelihood of each model via bridge sampling [@gronau2017bridge] and derive a Bayes factor. This requires `save_pars(all = TRUE)` during fitting (set in the `bmm()` call above):

```{r}
#| label: t1-bf-comparison
#| eval: false
bridge_softmax <- bridge_sampler(fit_m3_ss_softmax, repetitions = 10)
bridge_simple  <- bridge_sampler(fit_m3_ss_simple,  repetitions = 10)
bf(bridge_softmax, bridge_simple)
```

The Bayes factor strongly favors the softmax rule (*BF*~10~ = `r bf10_fmt`). This is consistent with recent comparisons by @oberauer2023measurement and @li2026filtering, who found softmax to outperform the simple rule across different working memory tasks. We use the softmax rule as the default in the remaining tutorials.

## Hypothesis Testing

Because we fitted the model with `sample_prior = "yes"`, we can use `brms::hypothesis()` to test directed hypotheses and obtain evidence ratios (Savage-Dickey density ratios comparing posterior to prior density at the hypothesis boundary). For example, to test whether context binding exceeds item memory within the open-set experiment:

```{r}
#| label: t1-hypothesis
#| eval: false
hypothesis(fit_m3_ss_softmax, "c_expopenset - a_expopenset > 0")
```

This returns an evidence ratio quantifying how much the data shift belief toward $c > a$ relative to the prior (ER = `r er_ca_fmt` for the open-set experiment). We can similarly test whether set size reduces $c$ (averaged across experiments) and whether set size effects differ between experiments:

```{r}
#| label: t1-hypothesis-setsize
#| eval: false
# Set size effect on context binding (averaged across experiments)
hypothesis(fit_m3_ss_softmax,
  "(c_expopenset:ss_lin + c_expclosedset:ss_lin) / 2 < 0")

# Experiment difference in set size effect on c
hypothesis(fit_m3_ss_softmax,
  "c_expopenset:ss_lin - c_expclosedset:ss_lin = 0")
```

The results confirm that $c > a$ in both experiments, that set size reduces both $a$ and $c$, and that the set size effect on $c$ is steeper for the closed-set experiment. The companion script contains the full set of six hypothesis tests. For other examples of $M^3$-based hypothesis testing with simple span data, see @loaiza2024active; for an application of the cs version to WM load and intelligence, see @schubert2023working.


# Tutorial 2: Complex Span $M^3$

This tutorial extends the workflow to tasks where distractors appear alongside memoranda. We use data from @li2026filtering, who developed a complex-span filtering task to test whether processed information can be kept out of working memory. The companion script is `scripts/tutorial2_complex_span.R`; data preparation is in `scripts/prepare_Li2026_data.R`.

## What makes complex span different

The simple span model has three response categories. When distractors are present, two more categories are needed: the distractor paired with the tested target (*paired distractor*) and distractors from other positions (*other distractor*). The complex span (cs) model extends the activation formulas with a filtering parameter $f$ that scales the activation of distractor categories:

$$
\begin{aligned}
\text{act}_{\text{correct}} &= b + a + c \\
\text{act}_{\text{paired dist.}} &= b + f \cdot a + f \cdot c \\
\text{act}_{\text{other}} &= b + a \\
\text{act}_{\text{other dist.}} &= b + f \cdot a \\
\text{act}_{\text{NPL}} &= b
\end{aligned}
$$

The parameter $f$ ranges from 0 to 1 (estimated on the logit scale). When $f = 0$, distractors receive only baseline activation (perfect filtering). When $f = 1$, distractors are activated as strongly as memoranda (no filtering).

A critical implementation detail: the `cs` version assigns activation formulas by position, so `resp_cats` must follow the order `corr, distc, other, disto, npl`. Misordering silently assigns the wrong formula to each category.

## Data

@li2026filtering presented participants with image pairs and manipulated whether targets and distractors were identified before processing (pre-cue), after processing (retro-cue), or whether no distractors were present (control). We use Experiment 1 (45 participants after exclusions, within-subjects). The data are pre-aggregated into response frequencies across the five categories; see the companion preparation script for classification details.

```{r}
#| label: t2-data-load
data_agg <- read_csv(here("data", "Li_2026_ComplexSpan_Exp1_agg.csv"),
                     show_col_types = FALSE) %>%
  mutate(condition = factor(condition, levels = c("control", "pre", "retro")))
```

## Model specification

The model definition follows the same pattern as Tutorial 1:

```{r}
#| label: t2-model-spec
m3_model_cs <- m3(
  resp_cats   = c("corr", "distc", "other", "disto", "npl"),
  num_options = c("n_corr", "n_distc", "n_other", "n_disto", "n_npl"),
  version     = "cs",
  choice_rule = "softmax"
)
```

The predictor formula specifies condition effects on all three parameters with cell means coding and uncorrelated random effects per condition:

```{r}
#| label: t2-formula
m3_formula_cs <- bmf(
  c ~ 0 + condition + (0 + condition || participant),
  a ~ 0 + condition + (0 + condition || participant),
  f ~ 0 + condition + (0 + condition || participant)
)
```

There is a complication: in the control condition, no distractors are present (`n_distc = 0`, `n_disto = 0`), so $f$ does not influence the likelihood and is not identified. If left free, the sampler will wander through the prior, and the resulting large variance can affect other parameters. The solution is to fix $f$ in the control condition using a `constant()` prior on both the fixed effect and the random effect SD:

```{r}
#| label: t2-priors
priors_cs <- c(
  prior(constant(10), nlpar = "f", coef = "conditioncontrol"),
  prior(constant(0),  class = "sd", nlpar = "f",
        coef = "conditioncontrol", group = "participant")
)
```

Setting the fixed effect to 10 on the logit scale (effectively $f \approx 1$) and the SD to 0 removes $f$ in the control condition from the model entirely.

## Model fitting and evaluation

Fitting follows the same `bmm()` call as in Tutorial 1, passing the custom priors. The model converged without issues ($\hat{R} \leq$ `r val_rhat_t2`, bulk ESS $>$ `r val_ess_t2`, no divergent transitions). @fig-t2-pp-check shows the posterior predictive check. The model captures the observed response proportions well across all three conditions and five categories.

```{r}
#| label: fig-t2-pp-check
#| fig-cap: "Posterior predictive check for the complex span $M^3$. Points show mean response proportions per category and condition; error bars show $\\pm 1$ SE across participants."
#| fig-alt: "Bar or point plot comparing observed and predicted response proportions for five categories (correct, paired distractor, other, other distractor, NPL) across control, pre-cue, and retro-cue conditions. Predicted values closely match observed proportions."
knitr::include_graphics(here("figures", "tutorial2_pp_check.pdf"))
```

## Parameter estimates and interpreting f

@fig-t2-param-estimates shows the posterior estimates. Item memory ($a$) and context binding ($c$) are on the identity link and can be read directly. The filtering parameter $f$ is estimated on the logit scale; to interpret it as a probability, we apply the inverse logit transformation (`plogis()` in R):

```{r}
#| label: t2-f-transform
#| eval: false
draws <- as_draws_df(fit_m3_cs) %>% as_tibble()
f_pre   <- plogis(draws$b_f_conditionpre)
f_retro <- plogis(draws$b_f_conditionretro)
```

```{r}
#| label: fig-t2-param-estimates
#| fig-cap: "Posterior parameter estimates for the complex span $M^3$. Left panels show item memory ($a$) and context binding ($c$) across conditions. Right panel shows distractor filtering ($f$) for pre-cue and retro-cue conditions, transformed to the probability scale. The control condition is omitted from the $f$ panel because it was fixed (no distractors present)."
#| fig-alt: "Multi-panel figure. Left panels show point-range plots of item memory (a) and context binding (c) across control, pre-cue, and retro-cue conditions. Right panel shows filtering parameter (f) on the probability scale for pre-cue and retro-cue, both near 1."
knitr::include_graphics(here("figures", "tutorial2_param_estimates.pdf"))
```

On the probability scale, $f$ is close to 1 in both conditions (pre-cue: $M$ = `r val_f_pre_m`, 95% HDI [`r val_f_pre_lo`, `r val_f_pre_hi`]; retro-cue: $M$ = `r val_f_retro_m`, 95% HDI [`r val_f_retro_lo`, `r val_f_retro_hi`]), indicating little to no filtering. This matches the conclusion of @li2026filtering: once distractors are processed, they are encoded into working memory regardless of whether participants knew in advance which items were distractors.

## Hypothesis tests

We can test condition differences formally using `brms::hypothesis()`:

```{r}
#| label: t2-hypothesis
#| eval: false
# Does f differ between pre-cue and retro-cue?
hypothesis(fit_m3_cs, "f_conditionpre - f_conditionretro = 0")

# Do a and c differ between control and distractor conditions?
hypothesis(fit_m3_cs,
  "a_conditioncontrol - (a_conditionpre + a_conditionretro) / 2 = 0")
```

The evidence ratio for the $f$ comparison is close to 1 (ER = `r er_f_fmt`), providing no evidence for a difference in filtering between cue conditions. Item memory and context binding are slightly lower in the distractor conditions than in the control, in line with a general cost of distractor processing on memory strength.

## Comparison with the published model

The standard `cs` model constrains distractor attenuation to be equal for item memory and context binding: both are scaled by the single parameter $f$. The original analysis by @li2026filtering used a custom model with two separate ratio parameters, $r_a$ and $r_c$, allowing item and binding activation to transfer differently to distractors. The `cs` model implicitly assumes $r_a = r_c = f$.

When this equal-attenuation constraint is too restrictive for the research question, a custom model is needed. Tutorial 3 demonstrates how to define such models from scratch.


# Tutorial 3: Custom $M^3$ — Parameter Recovery

The previous tutorials used pre-built model versions (`ss` and `cs`). When a task design does not fit one of these templates, `bmm` allows defining a fully custom $M^3$ with user-specified response categories, activation formulas, and link functions. This tutorial demonstrates the custom version through a parameter recovery simulation: we define a custom model, simulate data with known parameters, fit the model, and check whether the generating values are recovered. The walkthrough follows `scripts/tutorial3_parameter_recovery_simple.R`; design-grid results come from `scripts/tutorial3_parameter_recovery.R`.

## Why parameter recovery

Before applying a new or modified $M^3$ to empirical data, the model should be able to recover the parameters that generated the data. If recovery fails under ideal conditions (known generating values, no model misspecification), the problem lies in the model or the design, not in the data. Recovery simulations are especially informative for custom models where the number of parameters and the functional form of the activation equations are chosen by the researcher [for a systematic treatment, see @goettmann2025recovery; for the broader rationale of simulation-based validation in Bayesian workflows, see @schad2021toward and @wilson2019ten].

## Defining a custom $M^3$

As a running example, we use a memory updating task from @oberauer2019simple [applied with a custom $M^3$ by @li2025updating]. Participants memorize items at five positions, then each position is updated once (the old item is replaced). At test, one position is probed and the participant selects from a recognition set containing: the current target (1 option), other current items (4), the old item from the tested position (1), old items from other positions (4), and not-presented lures (5). This produces five response categories with the following activation formulas:

$$
\begin{aligned}
\text{act}_{\text{correct}}  &= b + (1 + e \cdot t_e) \cdot (a + c) \\
\text{act}_{\text{other}}    &= b + (1 + e \cdot t_e) \cdot a \\
\text{act}_{\text{oldinpos}} &= b + \exp(-r \cdot t_r) \cdot d \cdot (1 + e \cdot t_e) \cdot (a + c) \\
\text{act}_{\text{otherold}} &= b + \exp(-r \cdot t_r) \cdot d \cdot (1 + e \cdot t_e) \cdot a \\
\text{act}_{\text{NPL}}      &= b
\end{aligned}
$$

Beyond the familiar $a$ (item memory) and $c$ (context binding), three new parameters appear: $d$ captures the residual activation of replaced (outdated) items, $e$ scales the benefit of extended encoding time $t_e$, and $r$ governs how quickly old-item activation decays with removal time $t_r$. The design variables $t_e$ and $t_r$ are experimentally manipulated (each at 0.25, 0.5, and 1.0 s), producing a $3 \times 3$ within-subjects design with 9 conditions.

The model definition in `bmm` specifies the response categories, option counts, and link functions. Link functions are chosen to match each parameter's natural scale: identity for $a$ and $c$ (unbounded activation), logit for $d$ (bounded between 0 and 1), and log for $e$ and $r$ (positive values):

```{r}
#| label: t3-model-spec
m3_model_custom <- m3(
  resp_cats = c("correct", "other", "oldinpos", "otherold", "npl"),
  num_options = c(
    "n_correct" = 1, "n_other" = 4, "n_oldinpos" = 1,
    "n_otherold" = 4, "n_npl" = 5
  ),
  version     = "custom",
  choice_rule = "softmax",
  links = list(
    a = "identity", c = "identity",
    d = "logit", e = "log", r = "log"
  )
)
```

The activation formulas are specified separately using `bmf()`. Note that `te` and `tr` are data variables (not estimated parameters); `bmm` parses these expressions and passes them to `brms` as non-linear formulas:

```{r}
#| label: t3-activation-formulas
m3_act_funs <- bmf(
  correct  ~ (1 + e * te) * (a + c) + b,
  other    ~ (1 + e * te) * a + b,
  oldinpos ~ exp(-r * tr) * d * (1 + e * te) * (a + c) + b,
  otherold ~ exp(-r * tr) * d * (1 + e * te) * a + b,
  npl      ~ b
)
```

## Choosing generating parameter values

For the recovery simulation, we need group-level means and between-person standard deviations for each parameter. These are specified on the link scale (the scale on which the sampler operates), not the native scale. @tbl-gen-pars lists the generating values. The native range column shows the approximate span of individual differences ($\pm 2$ SD around the mean, transformed to native scale via the inverse link functions).

| Parameter | Link | Mean (link) | SD (link) | Native range ($\pm 2$ SD) |
|-----------|------|-------------|-----------|---------------------------|
| $c$ | identity | 2.0 | 0.50 | 1.0 -- 3.0 |
| $a$ | identity | 1.0 | 0.50 | 0.0 -- 2.0 |
| $d$ | logit | 0.85 | 0.65 | 0.38 -- 0.89 |
| $e$ | log | 0.00 | 0.35 | 0.50 -- 2.01 |
| $r$ | log | $-$0.69 | 0.38 | 0.24 -- 1.04 |

: Generating parameter values for the recovery simulation. All values are on the link scale. {#tbl-gen-pars}

Individual parameters are drawn from $\mathcal{N}(\mu_\text{link}, \sigma_\text{link})$ independently for each person and parameter. We simulate $N = 50$ participants, each completing 25 trials per condition (225 trials total). With set size 5, this corresponds to roughly one hour of testing.

## Simulating data with rm3()

The `rm3()` function generates multinomial response counts from a custom $M^3$ specification. It takes a vector of parameter values (on the native scale), the model definition, the activation formulas, and the design variables as inputs. Here is a call for a single participant in one condition:

```{r}
#| label: t3-rm3-demo
#| eval: false
demo_data <- rm3(
  n = 1, size = 25,
  pars = c(a = 1.2, c = 2.3, d = 0.7, e = 1.1, r = 0.5),
  m3_model = m3_model_custom,
  act_funs = m3_act_funs,
  te = 0.25, tr = 0.25
)
```

This returns a one-row data frame with response counts across the five categories. The full simulation loops over all 50 participants and 9 conditions, generating 450 rows of simulated data. The companion script contains the complete loop and data inspection code.

## Model fitting

We fit the simulated data with the same `bmm()` workflow used in the previous tutorials, now passing the custom model object:

```{r}
#| label: t3-fit
#| eval: false
# Prediction formula: intercept + random intercept per participant
m3_pred_funs <- bmf(
  correct  ~ (1 + e * te) * (a + c) + b,
  other    ~ (1 + e * te) * a + b,
  oldinpos ~ exp(-r * tr) * d * (1 + e * te) * (a + c) + b,
  otherold ~ exp(-r * tr) * d * (1 + e * te) * a + b,
  npl      ~ b,
  a ~ 1 + (1 || id), c ~ 1 + (1 || id),
  d ~ 1 + (1 || id), e ~ 1 + (1 || id), r ~ 1 + (1 || id)
)

fit <- bmm(
  formula = m3_pred_funs,
  model   = m3_model_custom,
  data    = sim_data,
  cores   = 4,
  backend = "cmdstanr",
  file    = here("output", "fit_m3_custom_simple_N50_T25")
)
```

Note that the prediction formula repeats the activation equations and adds regression formulas for each parameter (here, an intercept plus a random intercept per participant). Fitting a custom model with 5 parameters and 50 participants takes roughly 10--30 minutes depending on hardware.

## Recovery evaluation

### Group-level recovery

We first compare the recovered fixed effects to the true generating means. For each parameter, we extract the posterior median and 95% HDI from `fixef()` and check two things: (1) bias (is the posterior median close to the true value?) and (2) coverage (does the 95% HDI contain the true value?). We also recover the between-person SDs from `VarCorr()` and compare them to the generating SDs.

### Individual-level recovery

Individual-level recovery asks a finer question: for each participant, does the recovered estimate (fixed + random effect, extracted via `coef()`) match their true generating value? @fig-t3-indiv-scatter shows scatter plots of true versus recovered parameters for the single simulation cell ($N = 50$, 25 trials per condition). Points close to the diagonal indicate good recovery; the Pearson correlation ($r$) is annotated in each panel.

```{r}
#| label: fig-t3-indiv-scatter
#| fig-cap: "Individual-level parameter recovery for the single simulation cell ($N = 50$, 25 trials per condition). Each point is one participant; the dashed line indicates perfect recovery. Pearson correlations are shown per parameter."
#| fig-alt: "Five scatter plots, one per parameter (a, c, d, e, r), showing true generating values on the x-axis and recovered posterior medians on the y-axis. Points for a, c, and d cluster near the diagonal; points for e and r show more scatter."
knitr::include_graphics(here("figures", "tutorial3_simple_indiv_scatter.pdf"))
```

Parameters $a$, $c$, and $d$ show strong individual-level recovery ($r$ = `r val_cor_a`, `r val_cor_c`, and `r val_cor_d`, respectively). The parameters $e$ and $r$ are harder to recover ($r$ = `r val_cor_e` and `r val_cor_r`) because they interact multiplicatively with the design variables $t_e$ and $t_r$, which partially confounds them.

## Practical guidance

The simplified script demonstrates the workflow for a single design cell. To understand how recovery depends on sample size and the number of trials per condition, the full simulation script (`tutorial3_parameter_recovery.R`) varies both factors across a $3 \times 3$ grid: $N \in \{25, 50, 100\}$ and trials per condition $\in \{5, 10, 25\}$.

@fig-t3-group-recovery shows group-level recovery across the design grid. Recovered estimates are close to the true values in most cells, with HDIs narrowing as $N$ increases. With only 5 trials per condition, some parameters (particularly $e$ and $r$) show noticeable bias, but this improves substantially at 10 and 25 trials.

```{r}
#| label: fig-t3-group-recovery
#| fig-cap: "Group-level parameter recovery across the simulation design grid ($3 \\times 3$: sample sizes $\\times$ trials per condition). Points show recovered posterior medians; error bars show 95\\% HDIs. Dashed lines indicate the true generating values."
#| fig-alt: "Grid of point-range plots showing recovered group-level parameter estimates across 9 design cells (3 sample sizes by 3 trial counts). Dashed horizontal lines mark true values. HDIs narrow with larger N and more trials; some bias visible at 5 trials per condition."
knitr::include_graphics(here("figures", "tutorial3_group_recovery.pdf"))
```

@fig-t3-heatmap shows individual-level recovery (Pearson correlation between true and recovered values) for each parameter across the same design grid. Individual-level recovery is more demanding than group-level recovery: $e$ and $r$ require both a reasonable sample size and a sufficient number of trials per condition to reach correlations above .7.

```{r}
#| label: fig-t3-heatmap
#| fig-cap: "Individual-level parameter recovery (Pearson $r$) across the simulation design grid. Each panel shows one parameter; rows are sample sizes and columns are trials per condition. Higher correlations (darker shading) indicate better recovery of individual differences."
#| fig-alt: "Heatmap with five panels (one per parameter). Cells show Pearson correlations between true and recovered individual parameters across 9 design cells. Parameters a, c, and d show high correlations; e and r require more trials and larger samples."
knitr::include_graphics(here("figures", "tutorial3_recovery_heatmap.pdf"))
```

These results are based on a single replication per design cell. A proper simulation study should replicate each cell many times to account for Monte Carlo variability [see @goettmann2025recovery for a systematic approach, @chalmers2020simdesign for implementation tools, and @talts2018validating for a complementary calibration-based approach]. Recovery quality also depends on the model structure itself. Parameters that appear in interaction terms or that are only weakly identified by the design may require more data. Running a recovery check before collecting empirical data helps determine whether the planned sample size and number of trials are sufficient.


# Summary and Additional Resources

This tutorial covered three progressively complex applications of the $M^3$ framework using the `bmm` R package. The simple span example (Tutorial 1) introduced the full workflow: specification, fitting, evaluation, and hypothesis testing. It also compared the two choice rules. The complex span example (Tutorial 2) added distractor categories, the filtering parameter $f$, and the handling of non-identified parameters. The custom model example (Tutorial 3) showed how to define an $M^3$ from scratch for a memory updating task and how to validate it through parameter recovery.

A few practical recommendations. If the task produces three response categories (correct, other-list intrusion, NPL), the simple span version (`ss`) is the natural starting point. Tasks with distractors that are the same material type as memoranda fit the complex span version (`cs`). For other designs, the custom version gives full control over activation formulas, parameters, and link functions. In all cases, the softmax choice rule is the recommended default based on current evidence. When defining a custom model, running a parameter recovery simulation before data collection reveals whether the planned design can distinguish the parameters of interest and whether the model is identified.

## Additional resources

-   `bmm` package documentation and $M^3$ vignette: <https://venpopov.github.io/bmm/articles/bmm_m3.html>
-   Full companion scripts for all three tutorials: [repository link]
-   $M^3$ framework paper: @oberauer2019simple
-   Systematic parameter recovery for $M^3$: @goettmann2025recovery
-   Companion `bmm` tutorial for continuous-response mixture models: @frischkorn2025tutorial

<!-- ====================================================================== -->

# References {.unnumbered}

<!-- ====================================================================== -->

::: {#refs}
:::
