---
title: "Filtering Exp 1"
author: "Chenyu Li"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

# 1 Initial setting

## 1.1 Clear workspace and set default colors

```{r reset, include=FALSE}
# Clear all plots and variables from workspace
graphics.off()
rm(list=ls(all.names=TRUE))

# Set numerical precision and color schemes
options(digits = 3)
options(ggplot2.discrete.colour= c("#615F63","#FF7F6F","#2F7FC1","#FFBE7A","#8FC0A9","#8A1C56"))
options(ggplot2.discrete.fill= c("#615F63","#FF7F6F","#2F7FC1","#FFBE7A","#8FC0A9","#8A1C56"))
```

## 1.2 Import functions and packages

```{r}
# Install and load required packages
packages = c(
  # CRAN packages for data analysis and visualization
  "tidyverse", "brms", "ggpubr", "tidybayes", "emmeans", "bayestestR",
  # Github packages for specialized functions
  "chenyu-psy/smartr@v0.2.2", 
  "venpopov/bmm"
)
process_packages(packages, quietly = TRUE)

# Source all local R functions from the functions directory
load_functions("./functions")

# Set default ggplot theme
theme_set(theme_bw())

# Define position dodge constants for plotting
dodge1 = position_dodge(.1)
dodge2 = position_dodge(.2)

# Define task name and directory paths
task <- "Exp1"
model_path <- str_glue("./models/models_{task}/")
sample_path <- str_glue("./samples/sample_{task}/")
figure_path <- str_glue("./figures/figures_{task}/")

# Create directories if they don't exist
dir.create(file.path(model_path), showWarnings = FALSE)
dir.create(file.path(sample_path), showWarnings = FALSE)
dir.create(file.path(figure_path), showWarnings = FALSE)
```

# 2 Data processing

## 2.1 Import data

```{r}
# Load main experimental data
Raw_data <- str_glue("./data/Data_{task}.csv") %>% 
  read_csv() %>% 
  distinct()  # Remove any duplicate entries

# Load survey data (contains information about participant behavior)
Survey_data <- str_glue("./data/Survey_{task}.csv") %>% 
  read_csv()
```

## 2.2 Exclude participants based on survey responses

```{r}
# Identify participants who were distracted or used aids
excluding_aid_dist <- Survey_data %>% 
  filter(str_detect(aid, "Yes") | str_detect(attention, "No")) %>% 
  pull(participant)

message(str_glue("Excluded {length(excluding_aid_dist)} participants due to distraction or using aids."))
```

## 2.3 Analyze correlation between tasks

```{r}
# Calculate accuracy in the memory task for each participant
Acc_memory_task <- Raw_data %>% 
  filter(screenID == "retrieval", !participant %in% excluding_aid_dist) %>% 
  summarise(
    memory_acc = mean(acc, na.rm = TRUE),
    .by = c(participant)
  )

# Calculate accuracy in the judgment task for each participant
Acc_judge_task <- Raw_data %>% 
  filter(screenID == "memory", !participant %in% excluding_aid_dist) %>% 
  summarise(
    judge_acc = mean(acc, na.rm = TRUE),
    .by = c(participant)
  )

# Join the two datasets to analyze correlation
Acc_combine <- Acc_memory_task %>% 
  left_join(Acc_judge_task)

# Report correlation between tasks
message(paste0(
  "The correlation between the memory task and the size-judgement task is ",
  round(cor(Acc_combine["memory_acc"], Acc_combine["judge_acc"]), 2)))
```

## 2.4 Exclude low-performing participants

```{r}
# Identify participants with low accuracy in the size judgment task
excluding_lowAcc <- Raw_data %>% 
  filter(!participant %in% excluding_aid_dist) %>% 
  filter(screenID == "memory") %>% 
  mutate(
    # Handle NULL responses
    response = ifelse(response == "NULL", NA, response),
    # Calculate accuracy
    acc = case_when(
      is.na(response) ~ NA,
      acc ~ 1,
      TRUE ~ 0)
    ) %>% 
  summarise(
    pResp = sum(!is.na(response), na.rm = TRUE)/n(),  # Response rate
    pAcc = mean(acc, na.rm = TRUE),                   # Accuracy rate
    .by = participant
  ) %>% 
  filter((pResp < 0.5 | pAcc < 0.7)) %>%              # Exclusion criteria
  pull(participant)

message(str_glue("Excluded {length(excluding_lowAcc)} participants due to a low accuracy in the size judgement task"))

```

## 2.5 Clean data and remove trials with missing images

```{r}
# Process judgment data, identifying trials with missing images
Data_judge <- Raw_data %>% 
  # Apply exclusion criteria
  filter(!participant %in% c(excluding_aid_dist, excluding_lowAcc)) %>% 
  filter(screenID == "memory") %>% 
  mutate(
    # Mark trials with missing images
    markNA = ifelse(is.na(leftImage) | is.na(rightImage), 1, 0),
    # Identify target and distractor images
    target = ifelse(cue == "left", leftImage, rightImage),
    distractor = ifelse(cue == "left", rightImage, leftImage),
    # Convert RT to seconds
    RT = rt/1000
  ) %>% 
  # Count missing images per trial
  mutate(
    markNA = sum(markNA),
    .by = c("participant", "blockID", "trialID")
  ) %>% 
  # Rename accuracy column for clarity
  rename(judgeAcc = acc) %>% 
  # Select relevant columns
  select(
    participant, blockID, trialID, condition, congruency, target, distractor, judgeAcc, RT, markNA
  )

# Extract trials with missing images for reporting
Filtered_trials <- Data_judge %>% 
  filter(markNA > 0)

# Remove trials with missing images from the main dataset
Data_judge <- Data_judge %>% 
  filter(markNA == 0) %>% 
  select(-markNA)

message(str_glue("Excluded {n_distinct(Filtered_trials[,c('blockID','trialID')])} trials from {n_distinct(Filtered_trials$participant)} participants due to missing images"))
```

## 2.6 Process retrieval task data

```{r}
# Prepare retrieval data and merge with judgment data
Data_Ret <- Raw_data %>% 
  filter(screenID == "retrieval") %>% 
  select(participant, blockID, trialID, condition, congruency, correctObject, response) %>% 
  rename(target = correctObject) %>% 
  right_join(Data_judge) %>% 
  # Calculate response categories
  mutate(
    Correct = ifelse(target == response, 1, 0),                  # Correct responses
    ColorDist = ifelse(distractor == response, 1, 0),            # Color distractor responses
    noRes = ifelse(response == "NULL", 1, 0)                     # No response
  ) %>% 
  # Calculate additional response categories
  mutate(
    Other = ifelse(Correct == 0 & response %in% target, 1, 0),    # Other response with target
    OtherDist = ifelse(ColorDist == 0 & response %in% distractor, 1, 0), # Other response with distractor
    NPL = 1 - Correct - Other - ColorDist - OtherDist - noRes,    # Non-presented lures
    .by = c(participant, blockID, trialID)
  )
```

## 2.7 Calculate response options for each condition

```{r}
# Create table of available options for each response type by condition
Table_nOptions <- data.frame(
  condition = c("control", "pre", "retro")) %>% 
  mutate(
    nCorrect = 1,                                           # Always 1 correct answer
    nOther = 2,                                             # Always 2 other options
    nColorDist = ifelse(condition == "control", 0, 1),      # Color distractor options
    nOtherDist = ifelse(condition == "control", 0, 2),      # Other distractor options
    nNPL = ifelse(condition == "control", 9, 6)             # Non-presented lure options
  )

# Format option counts for later use
Columns_nOptions <- Table_nOptions %>% 
  pivot_longer(cols = starts_with("n"), names_to = "response", values_to = "nOpt") %>% 
  mutate(response = substring(response, 2))
```

# 3 Data Visualization

## 3.1 Visualize judgment task performance

```{r}
# Plot accuracy in the judgment task
Plot_juadg_Acc <- Data_judge %>% 
  # Aggregate data for plotting
  agg_plot(y = judgeAcc, within = c("condition"), group = participant) %>%
  # Add condition labels
  labelCondition() %>% 
  ggplot(aes(x = condition, y = mean, fill = condition)) +
  # Add error bars and points
  geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), width = .05) +
  geom_point(
    shape = 23,
    size = 3
  ) +
  ylim(0.5, 1) +
  # Add labels
  labs(
    y = "Proportion of correct", 
    x = "Condition", 
    color = "Congruency",
    title = "Accuracy",
    tag = "A"
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10.5)
  )

# Plot reaction time in the judgment task
Plot_judge_RT <- Data_judge %>% 
  # Only include correct responses
  filter(judgeAcc == 1) %>%
  # Aggregate data for plotting
  agg_plot(y = RT, within = c("condition"), group = participant) %>%
  # Add condition labels
  labelCondition() %>% 
  ggplot(aes(x = condition, y = mean, fill = condition)) +
  # Add error bars and points
  geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), width = .05) +
  geom_point(
    shape = 23,
    size = 3
  ) +
  ylim(1.2, 1.6) +
  # Add labels
  labs(
    y = "Reaction time (s)", 
    x = "Condition", 
    color = "Congruency",
    title = "Reaction time",
    tag = "B"
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10.5)
  )

# Combine accuracy and RT plots
(Plot_juadgement <- ggarrange(
  Plot_juadg_Acc, Plot_judge_RT,
  ncol = 2, nrow = 1, widths = c(1, 1),
  common.legend = FALSE)
)

# Save the combined plot
ggsave(
  filename = str_glue("{figure_path}Plot_{task}_judgement.pdf"),
  plot = Plot_juadgement,
  width = 6, height = 4
)
```

## 3.2 Visualize memory task performance

```{r}

Table_Res_NPL <- Data_Ret %>% 
  filter(!str_detect(condition, "control")) %>% 
  agg_plot(y = NPL, within = c("condition"), group = participant) %>% 
  mutate(
    same = mean/6,
    other = mean/6*2
  ) %>% 
  select(condition, same, other) %>%
  pivot_longer(cols = c("same", "other"), names_to = "distractor", values_to = "value") %>% 
  labelCondition() %>% 
  mutate(distractor = relabel(distractor, labels = c(
        "same" = "Color distractor",
        "other" = "Other distractors"
      )))


(
  Plot_Resp_correct <- Data_Ret %>% 
    agg_plot(y = Correct, within = c("condition"), group = participant) %>%
    labelCondition() %>%
    ggplot(aes(x = condition, y = mean, fill=condition)) +
    geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), width = .05)+
    geom_point(
      shape = 23,
      size = 3
    )+
    ylim(0.5,1)+
    labs(
      title = "Correct",
      tag = "A",
      y = "Proportion of correct", 
      x = "Condition")+
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10.5)
  )
)

(
  Plot_Resp_dist <- Data_Ret %>%
    select(participant, blockID, trialID, condition, congruency, ColorDist, OtherDist) %>%
    filter(condition != "control") %>%
    pivot_longer(
      cols = c("OtherDist", "ColorDist"), 
      names_to = "distractor") %>%
    mutate(distractor = relabel(distractor, labels = c(
      "ColorDist" = "Color distractor", 
      "OtherDist" = "Other distractors"
    ))) %>% 
    agg_plot(y = value, within = c("condition"), between="distractor", group = participant) %>%
    labelCondition() %>%
    ggplot(aes(x = condition, y = mean, fill = condition))+
    geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), width = .05)+
    geom_point(
      shape = 23,
      size = 3
    )+
    geom_line(
      data = Table_Res_NPL,
      mapping = aes(
        x = condition,
        y = value,
        group = 1
      ),
      linetype = "dashed",
      linewidth = 1
    )+
    geom_point(
      data = Table_Res_NPL,
      mapping = aes(
        x = condition,
        y = value
      ),
      shape = 23,
      size = 3,
    )+
    scale_y_continuous(
      limits = c(0,0.12),
      breaks = c(0,0.03,0.06,0.09,0.12))+
    scale_fill_manual(values = c("#FF7F6F","#2F7FC1"))+
    labs(
      title = "Distractor",
      tag = "B",
      y = "Proportion of response", 
      x = "Condition")+
    facet_wrap(~distractor)+
    theme(
      legend.position = "none",
      axis.text = element_text(size = 10.5),
      strip.text = element_text(size = 10.5),
      strip.background = element_rect(fill = "#bababa")
    )
)

Plot_response <- ggarrange(
  Plot_Resp_correct,NA, Plot_Resp_dist,
  ncol = 3, nrow = 1, widths = c(1.5,0.3, 2),
  common.legend = FALSE
)

ggsave(
  filename = str_glue("{figure_path}Plot_{task}_response.pdf"),
  plot = Plot_response,
  width = 10, height = 4)

```



# 4 Model fitting

## 4.1 Size judgement task

### 4.1.1 Accuracy

```{r}

mData_judge_correct <- Data_judge %>% 
  summarise(
    score = sum(judgeAcc),
    nRet = n(),
    .by = c(participant, condition)
    ) %>% 
  mutate(
    condition = factor(condition)
  )

# 
contrasts(mData_judge_correct$condition) = bayestestR::contr.equalprior(n = 3)

smart_runFun(
  fun = brm,
  args = list(
    formula = score | trials(nRet) ~ 1 + condition + (1 + condition || participant),
    data = mData_judge_correct,
    family = binomial(),
    prior = c(prior(normal(0,1), class = "b"), 
              prior(student_t(3, 0, 2), class = "sd", lb = 0)),
    chains = 8,
    iter = 10000,
    warmup = 2500,
    cores = 8,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    save_pars = save_pars(all = TRUE),
    file = str_glue("{model_path}Model_{task}_judge_Correct")
  ),
  name = str_glue("Model_{task}_judge_Correct"),
)

Model_judge_Correct <- readRDS(str_glue("{model_path}Model_{task}_judge_Correct.rds"))

smart_runFun(
  fun = unupdate,
  args = list(model = Model_judge_Correct),
  name = str_glue("Prior_{task}_judge_Correct"),
  export = "Prior_judge_Correct"
)

# The set-size effect
pairwise_comparisons(Model_judge_Correct, Prior_judge_Correct, specs = "~ condition")

# Parameter       |       BF
# --------------------------
# control - pre   | 1.46e+03
# control - retro | 2.75e+03
# pre - retro     |    0.071



```

### 4.1.2 Reaction time

```{r}
mData_judge_RT <- Data_judge %>% 
  filter(judgeAcc==1) %>% 
  mutate(condition = factor(condition))

contrasts(mData_judge_RT$condition) = bayestestR::contr.equalprior(n = 3)

smart_runFun(
  fun = brm,
  args = list(
    formula = RT ~ 1 + condition + (1 + condition || participant),
    data = mData_judge_RT,
    family = lognormal(),
    prior = c(prior(cauchy(0, sqrt(2)/4), class = "b"), 
              prior(student_t(3, 0, 2), class = "sd", lb = 0)),
    chains = 8,
    iter = 10000,
    warmup = 2500,
    cores = 8,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    save_pars = save_pars(all = TRUE),
    file = str_glue("{model_path}Model_{task}_judge_RT")
  ),
  name = str_glue("Model_{task}_judge_RT"),
)


Model_judge_RT <- readRDS(str_glue("{model_path}Model_{task}_judge_RT.rds"))

smart_runFun(
  fun = unupdate,
  args = list(model = Model_judge_RT),
  name = str_glue("Prior_{task}_judge_RT"),
  export = "Prior_judge_RT"
)


# The set-size effect
pairwise_comparisons(Model_judge_RT, Prior_judge_RT, specs = "~ condition")

# Parameter       |    BF
# -----------------------
# control - pre   |  2.17
# control - retro |  6.10
# pre - retro     | 0.020


```

## 4.2 Memory task

### 4.2.1 Data

```{r}

mData_Ret <- Data_Ret %>% 
  select(participant,condition,Correct,ColorDist,OtherDist,NPL) %>% 
  pivot_longer(
    cols = c(Correct,ColorDist, OtherDist, NPL),
    names_to = "response",values_to = "value") %>% 
  summarise(
    Resp = sum(value, na.rm = TRUE),
    nResp = sum(!is.na(value)),
    .by = c(participant, condition, response)
  ) %>% 
  mutate(
    response = factor(response),
    condition = factor(condition)
  )

mArgs_Ret = list(
    formula = Resp | trials(nResp) ~ 1 + condition + (1 + condition || participant),
    family = binomial(),
    prior = c(prior(normal(0,1), class = "b"), 
              prior(student_t(3, 0, 2), class = "sd", lb = 0)),
    chains = 8,
    iter = 10000,
    warmup = 2500,
    cores = 8,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    save_pars = save_pars(all = TRUE),
    sample_prior = "yes"
  )

```


### 4.2.2 Response: Correct

```{r}

mData_Ret_correct <- mData_Ret %>% 
  filter(response == "Correct")

contrasts(mData_Ret_correct$condition) = bayestestR::contr.equalprior(n = 3)

smart_runFun(
  fun = brm,
  args = mArgs_Ret %>% 
    list_assign(
      data = mData_Ret_correct,
      file = str_glue("{model_path}Model_{task}_memory_Correct")
    ),
  name = str_glue("Model_{task}_memory_Correct"),
)

Model_memory_Correct <- readRDS(str_glue("{model_path}Model_{task}_memory_Correct.rds"))

smart_runFun(
  fun = unupdate,
  args = list(model = Model_memory_Correct),
  name = str_glue("Prior_{task}_memory_Correct"),
  export = "Prior_memory_Correct"
)


pairwise_comparisons(Model_memory_Correct, Prior_memory_Correct, specs = "~ condition")

# Parameter       |       BF
# --------------------------
# control - pre   | 6.68e+09
# control - retro | 2.81e+08
# pre - retro     |    0.061


```

### 4.2.3 Response: Distractor

```{r}

mData_Ret_dist <- mData_Ret %>% 
  filter(response=="ColorDist",condition!="control") %>% 
  droplevels()

contrasts(mData_Ret_dist$condition) = bayestestR::contr.equalprior(n = 2)

smart_runFun(
  fun = brm,
  args = mArgs_Ret %>% 
    list_assign(
      data = mData_Ret_dist,
      file = str_glue("{model_path}Model_{task}_memory_ColorDist")
    ),
  name = str_glue("Model_{task}_memory_ColorDist"),
)

mData_Ret_otherDist <- mData_Ret %>% 
  filter(response=="OtherDist",condition!="control") %>% 
  droplevels()

contrasts(mData_Ret_otherDist$condition) = bayestestR::contr.equalprior(n = 2)

smart_runFun(
  fun = brm,
   args = mArgs_Ret %>% 
    list_assign(
      data = mData_Ret_otherDist,
      file = str_glue("{model_path}Model_{task}_memory_otherDist")
    ),
  name = str_glue("Model_{task}_memory_OtherDist"),
)


Model_memory_ColorDist <- readRDS(str_glue("{model_path}Model_{task}_memory_ColorDist.rds"))
Model_memory_OtherDist <- readRDS(str_glue("{model_path}Model_{task}_memory_OtherDist.rds"))

smart_runFun(
  fun = unupdate,
  args = list(model = Model_memory_ColorDist),
  name = str_glue("Prior_{task}_memory_ColorDist"),
  export = "Prior_memory_ColorDist"
)

smart_runFun(
  fun = unupdate,
  args = list(model = Model_memory_OtherDist),
  name = str_glue("Prior_{task}_memory_OtherDist"),
  export = "Prior_memory_OtherDist"
)


pairwise_comparisons( Model_memory_ColorDist, Prior_memory_ColorDist, specs = "~ condition")

# Parameter   |    BF
# -------------------
# pre - retro | 0.282


pairwise_comparisons( Model_memory_OtherDist, Prior_memory_OtherDist, specs = "~ condition")

# Parameter   |    BF
# -------------------
# pre - retro | 0.087

```

### 4.2.4 ColorDist vs. NPL

```{r}

mData_Ret_dist_NPL <- mData_Ret %>% 
  filter(
    response %in% c("ColorDist", "NPL"),
    condition != "control"
  ) %>% 
  droplevels()

contrasts(mData_Ret_dist_NPL$condition) = bayestestR::contr.equalprior(n = 2)
contrasts(mData_Ret_dist_NPL$response) = bayestestR::contr.equalprior(n = 2)

smart_runFun(
  fun = brm,
  args = mArgs_Ret %>% 
    list_modify(
      formula = Resp | trials(nResp) ~ 0 + condition:response + (0 + condition:response || participant),
    ) %>% 
    list_assign(
      data = mData_Ret_dist_NPL,
      file = str_glue("{model_path}Model_{task}_ColorDist_NPL")
    ),
  name = str_glue("Model_{task}_ColorDist_NPL")
)

Model_Ret_dist_NPL <- readRDS(str_glue("{model_path}Model_{task}_ColorDist_NPL.rds"))

Hyp_Ret_dist_NPL <- Model_Ret_dist_NPL %>% 
  hypothesis(c(
    precue = "inv_logit(conditionpre:responsedist) = inv_logit(conditionpre:responseNPL)/6",
    retrocue = "inv_logit(conditionretro:responsedist) = inv_logit(conditionretro:responseNPL)/6"
  ))

Hyp_Ret_dist_NPL$hypothesis %>% mutate(Evid.Ratio = 1/Evid.Ratio)

#   Hypothesis Estimate Est.Error CI.Lower CI.Upper Evid.Ratio Post.Prob Star
# 1     precue   0.0867   0.00878   0.0698   0.1044   3.37e+16  2.97e-17    *
# 2   retrocue   0.0764   0.00672   0.0632   0.0896   2.28e+17  4.39e-18    *

```


### 4.2.5 OtherDist vs. NPL

```{r}

mData_Ret_otherDist_NPL <- mData_Ret %>% 
  filter(
    response %in% c("OtherDist", "NPL"),
    condition != "control"
  ) %>% 
  droplevels()

contrasts(mData_Ret_otherDist_NPL$condition) = bayestestR::contr.equalprior(n = 2)
contrasts(mData_Ret_otherDist_NPL$response) = bayestestR::contr.equalprior(n = 2)

smart_runFun(
  fun = brm,
  args = mArgs_Ret %>% 
    list_modify(
      formula = Resp | trials(nResp) ~ 0 + condition:response + (0 + condition:response || participant),
    ) %>% 
    list_assign(
      data = mData_Ret_otherDist_NPL,
      file = str_glue("{model_path}Model_{task}_OtherDist_NPL")
    ),
  name = str_glue("Model_{task}_OtherDist_NPL")
)

Model_Ret_otherDist_NPL <- readRDS(str_glue("{model_path}Model_{task}_OtherDist_NPL.rds"))

Hyp_Ret_otherDist_NPL <- Model_Ret_otherDist_NPL %>% 
  hypothesis(c(
    precue = "inv_logit(conditionpre:responseotherDist) = inv_logit(conditionpre:responseNPL)/6",
    retrocue = "inv_logit(conditionretro:responseotherDist) = inv_logit(conditionretro:responseNPL)/6"
  ))

Hyp_Ret_otherDist_NPL$hypothesis %>% mutate(Evid.Ratio = 1/Evid.Ratio)

#   Hypothesis Estimate Est.Error CI.Lower CI.Upper Evid.Ratio Post.Prob Star
# 1     precue   0.0744   0.00888   0.0578   0.0925   4.96e+17  2.02e-18    *
# 2   retrocue   0.0744   0.00921   0.0570   0.0933   1.87e+14  5.34e-15    *

```

## 4.3 Memory measurement models

### 4.3.1 Data Preparation

```{r}

mData_M3 <- Data_Ret %>% 
  agg_multinomial(
    responses = c("Correct", "Other", "ColorDist", "OtherDist", "NPL"),
    group = c("participant", "condition"),
    DV_name = NULL
  ) %>% 
  left_join(Table_nOptions) %>% 
  mutate(condition = factor(condition, levels = c("control","pre","retro")))

contrasts(mData_M3$condition) = bayestestR::contr.equalprior(n = 3)

write_rds(mData_M3, str_glue("./data/Data_{task}_M3.rds"))

```


### 4.3.2 formulas and priors

```{r}

M3_form = bmf(
  Correct ~ a + c + b,
  Other ~ a + b,
  ColorDist ~ ra*a + rc*c + b,
  OtherDist ~ ra*a + b,
  NPL ~ b,
  a ~ 0 + condition + (0 + condition || participant),
  c ~ 0 + condition + (0 + condition || participant),
  ra ~ 0 + condition + (0 + condition || participant),
  rc ~ 0 + condition + (0 + condition || participant)
)

M3_model <- m3(
    resp_cats = c("Correct","Other","ColorDist", "OtherDist","NPL"),
    num_options = c("nCorrect","nOther","nColorDist", "nOtherDist", "nNPL"),
    choice_rule = "softmax",
    links = list(
      a = "identity",
      c = "identity",
      ra = "logit",
      rc = "logit"),
    default_priors = list(
      a = list(main = "normal(3,1)", effect = "normal(3,1)"),
      c = list(main = "normal(3,1)", effect = "normal(3,1)"),
      ra = list(main = "logistic(0,1)", effect = "logistic(0,1)"),
      rc = list(main = "logistic(0,1)", effect = "logistic(0,1)"))
    )

M3_prior <- c(
      prior(constant(10), class="b", nlpar = "ra", coef = "conditioncontrol"),
      prior(constant(10), class="b", nlpar = "rc", coef = "conditioncontrol")
    )

```

### 4.3.3 Model fitting

```{r}

smart_runFun(
  fun = bmm,
  args = list(
    formula = M3_form,
    model = M3_model,
    data = mData_M3,
    prior = M3_prior,
    sample_prior = "yes",
    iter = 10000,
    warmup = 2500,
    chains = 8,
    cores = 8,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    save_pars = save_pars(all = TRUE),
    file = str_glue("{model_path}Model_{task}_M3")
  ),
  name = str_glue("Model_{task}_M3"),
)

```


### 4.3.4 Bayes Factor

```{r}

Model_M3 <- read_rds(str_glue("{model_path}Model_{task}_M3.rds"))
Model_M3_simple <- read_rds(str_glue("{model_path}Model_{task}_M3_simple.rds"))

brms::bayes_factor(Model_M3, Model_M3_simple)

Hyp_M3 <- hypothesis(Model_M3, c(
  a_pre_rertro = "a_conditionpre = a_conditionretro",
  a_pre_control = "a_conditionpre = a_conditioncontrol",
  a_retro_control = "a_conditionretro = a_conditioncontrol",

  c_pre_rertro = "c_conditionpre = c_conditionretro",
  c_pre_control = "c_conditionpre = c_conditioncontrol",
  c_retro_control = "c_conditionretro = c_conditioncontrol",
  
  ra_pre_rertro = "inv_logit(ra_conditionpre) = inv_logit(ra_conditionretro)",
  rc_pre_rertro = "inv_logit(rc_conditionpre) = inv_logit(rc_conditionretro)"
))

# Hypothesis Tests for class b:
#        Hypothesis Estimate Est.Error CI.Lower CI.Upper Evid.Ratio Post.Prob Star
# 1    a_pre_rertro     0.32      0.34    -0.33     1.00       2.67      0.73     
# 2   a_pre_control    -0.05      0.33    -0.68     0.61       4.40      0.81     
# 3 a_retro_control    -0.37      0.29    -0.94     0.20       2.07      0.67     

# 4    c_pre_rertro     0.05      0.22    -0.39     0.49       6.26      0.86     
# 5   c_pre_control    -0.37      0.21    -0.79     0.05       1.45      0.59     
# 6 c_retro_control    -0.42      0.21    -0.84    -0.01       0.89      0.47    *

# 7   ra_pre_rertro     0.04      0.04    -0.05     0.12       6.31      0.86     
# 8   rc_pre_rertro     0.03      0.07    -0.11     0.18       5.21      0.84     

```

### 4.3.5 Posterior

```{r}

Model_M3 <- read_rds(str_glue("{model_path}Model_{task}_M3.rds"))

Post_M3 <- Model_M3 %>% 
  gather_draws(`b_.*`, ndraws=8000, regex = TRUE) %>% 
  separate_wider_delim(.variable, "_", names = c(NA, "par","condition"), cols_remove = TRUE) %>%  
  mutate(
    condition = str_remove(condition,"condition")) %>% 
  rename(value = .value) %>% 
  mutate(value = ifelse(str_detect(par,"r"),inv_logit_scaled(value), value)) %>% 
  filter(!(condition=="control" & str_detect(par,"r")))

Post_M3 %>% write_csv(str_glue('./figures/figures_Data/Post_{task}.csv'))


```

### 4.3.6 Prediction

```{r}

Model_M3 <- read_rds(str_glue("{model_path}Model_{task}_M3.rds"))
inv_logit <- brms::inv_logit_scaled

Table_Res_filtered <- mData_M3 %>% 
  select(participant, condition, Correct, Other, ColorDist, OtherDist, NPL, nRet) %>% 
  pivot_longer(
    cols = c(Correct, Other, ColorDist, OtherDist, NPL), 
    names_to = ".category", values_to = ".epred") %>% 
  mutate(.epred = .epred/nRet,
         .category = relabel(.category, labels = c(
           "Correct" = "Correct",
           "ColorDist" = "Color distractor",
           "NPL" = "NPL",
           "Other" = "Others",
           "OtherDist" = "Other distractors"
         ))) %>% 
  labelCondition()

Pred_M3 <- Model_M3 %>% 
  epred_draws(
    newdata = Model_M3$data,
    ndraws = 10000
  ) %>% 
  ungroup() %>% 
  mutate(.epred = .epred/nTrials) %>% 
  select(participant, condition, .draw, .category, .epred) %>% 
  summarise(.epred = mean(.epred),
            .by = c(.draw, condition, .category)) %>% 
  filter(
    .epred > quantile(.epred, 0.005),
    .epred < quantile(.epred, 0.995),
    .by = c(condition, .category)
  ) %>% 
  filter(!(condition=="control" & .category %in% c("ColorDist","OtherDist"))) %>% 
  mutate(.category = relabel(.category, labels = c(
           "Correct" = "Correct",
           "ColorDist" = "Color distractor",
           "NPL" = "NPL",
           "Other" = "Others",
           "OtherDist" = "Other distractors"
         ))) %>% 
  labelCondition()

```


```{r}

(
  Plot_M3_recovery<- Table_Res_filtered %>% 
    agg_plot(y = .epred, within = c(".category","condition"), group = participant) %>% 
    ggplot(aes(x = condition, y = mean)) +
    geom_point(position = dodge1)+
    gghalves::geom_half_violin(
      mapping = aes(x = condition, y = .epred, fill = condition),
      data = filter(Pred_M3),
      side = "r", nudge = 0.1)+
    geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), width = 0, position = dodge1)+
    ylim(0,1)+
    labs(y = "Proportion of response", x = "Condition")+
    facet_wrap(~ .category, scales = "free")+
    theme(
      legend.position = "none",
      axis.text = element_text(size = 10.5),
      strip.text = element_text(size=10.5),
      strip.background = element_rect(fill = "#bababa"),
      panel.spacing.x = unit(3, "lines"),
      panel.spacing.y = unit(1, "lines") )
  )

scale_inidividual_facet_y_axes(
  Plot_M3_recovery,
  ylims = list(
    c(0.4,1), c(0, 0.30), c(0, 0.08), 
    c(0, 0.30), c(0, 0.30)
  )
)

ggsave(
  filename = str_glue("{figure_path}Plot_{task}_M3_predict.pdf"),
  plot = Plot_M3_recovery, width = 10, height =6
)

```

## 4.4 M3 from OberLewan 2019

### 4.4.1 Data Preparation

```{r}

mData_M3 <- Data_Ret %>% 
  filter(condition!="control") %>% 
  agg_multinomial(
    responses = c("Correct", "Other", "ColorDist", "OtherDist", "NPL"),
    group = c("participant","condition"),
    DV_name = NULL
  ) %>% 
  mutate(i = ifelse(condition=="pre",1,0)) %>% 
  select(-condition)


```

### 4.4.2 formulas and priors

```{r}

genM3Form <- function(choice=NULL, f = "both", r = "both") {
  
  if (f == "both") {
    fa = "pow(f,i)"
    fc = "pow(f,i)"
  } else if(f == "a") {
    fa = "pow(f,i)"
    fc = NULL
  } else if(f == "c") {
    fa = NULL
    fc = "pow(f,i)"
  } else {
    fa = NULL
    fc = NULL
  }
  
  if (r == "both") {
    ra = "r"
    rc = "r"
  } else if(r == "a") {
    ra = "r"
    rc = NULL
  } else if(r == "c") {
    ra = NULL
    rc = "r"
  } else {
    ra = NULL
    rc = NULL
  }
  
  
  dis_a = paste(c(fa,ra,"a"),collapse = '*')
  dis_c = paste(c(fc,rc,"c"),collapse = "*")
  form_dist = str_glue("ColorDist ~ {dis_a} + {dis_c} + b") %>% as.formula()
  form_otherDist = str_glue("OtherDist ~ {dis_a} + b") %>% as.formula()
  
  if (f!="None") {
    f_form = f ~ 1 + (1 || participant)
  } else {
    f_form = NULL
  }
  
  if (r!="None") {
    r_form = r ~ 1 + (1 || participant)
  } else {
    r_form = NULL
  }
  
    M3_form = bmf(
      Correct ~ a + c + b,
      Other ~ a + b,
      form_dist,
      form_otherDist,
      NPL ~ b,
      a ~ 1 + (1 || participant),
      c ~ 1 + (1 || participant))+
      f_form +
      r_form
  
  return(M3_form)
  
}

```

### 4.4.3 priors

```{r}

genM3Model <- function(choice="softmax", f = "both", r = "both") {
  
  if (choice=="softmax") {
    M3_list = list(
      a = "identity",
      c = "identity")
    
    M3_default_priors <- list(
        a = list(main = "normal(3,1)", effect = "normal(3,1)"),
        c = list(main = "normal(3,1)", effect = "normal(3,1)"))
    
  } else if (choice=="simple") {
    
    M3_list = list(
      a = "log",
      c = "log")
    
    M3_default_priors <- list(
      a = list(main = "normal(1.5,0.5)", effect = "normal(1.5,0.5)"),
      c = list(main = "normal(1.5,0.5)", effect = "normal(1.5,0.5)"))
    
  } else {
    
    stop("The choice argument supports only `softmax` or `simple`.")
  }
  
  
  if (f!="None") {
    M3_list["f"] = "logit"
    M3_default_priors <- M3_default_priors %>% list_assign(f = list(main = "logistic(0,1)", effect = "logistic(0,1)"))
  }
  
  if (r!="None") {
    M3_list["r"] = "logit"
    M3_default_priors <- M3_default_priors %>% list_assign(r = list(main = "logistic(0,1)", effect = "logistic(0,1)"))
  }
  
  M3_model <- m3(
    resp_cats = c("Correct","Other","ColorDist", "OtherDist","NPL"),
    num_options = c(1,2,1,2,6),
    choice_rule = choice,
    links = M3_list,
    default_priors = M3_default_priors
    )
  
  return(M3_model)
}

```


### 4.4.4 Model fitting

```{r}

MC_list <- list(
  choice = c("softmax"),
  r = c("both", "a", "c", "None"),
  f = c("both", "a", "c", "None")
)


smartr::parallel_model_comparison(
  fun = bmm,
  pars = MC_list,
  form_fun = genM3Form,
  model_fun = genM3Model,
  args = list(
    data = mData_M3,
    sample_prior = "yes",
    iter = 10000,
    warmup = 2500,
    chains = 16,
    cores = 4,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    save_pars = save_pars(all = TRUE)
  ),
  model_name = "Model_M3_OL2019",
  model_path = model_path,
  sample_path = sample_path,
  maxCore = 8
)

smart_runFun(
  fun = bmm,
  args = list(
    data = mData_M3,
    sample_prior = "yes",
    formula = genM3Form(choice = "softmax", f = "a", r = "both"),
    model = genM3Model(choice = "softmax", f = "a", r = "both"),
    iter = 10000,
    warmup = 2500,
    chains = 16,
    cores = 8,
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    save_pars = save_pars(all = TRUE),
    file = str_glue("{model_path}Model_M3_OL2019_choicesoftmax_rboth_fa.rds")
  ),
  name = str_glue("Model_M3_OL2019_choicesoftmax_rboth_fa")
)


```


### 4.4.5 Visualization

```{r}

 Table_compete <- compare_logml(
   pars = MC_list,
   sample_path = sample_path,
   sample_prefix = "Sample_M3_OL2019"
 ) %>% 
  mutate(
    logml = max(logml) - logml,
    logml = round(logml,2),
    f = relabel(f, labels = c(
        "both" = "a & c",
        "a" = "a",
        "c" = "c",
        "None" = "0"
      )),
    r = relabel(r, labels = c(
        "both" = "r & f",
        "a" = "r",
        "c" = "f",
        "None" = "0"
      ))
    )


Plot_logml_heatmap <- Table_compete %>% 
  ggplot(aes(x = f, y = r, fill = logml))+
  geom_tile()+
  geom_text(aes(label = logml))+
  scale_fill_gradient(
    low = "white", high = "#2F7FC1", 
    trans = scales::exp_trans(base = log(2.7)),
    breaks = c(0, 20, 50, 100, 1000)
    )+
  labs(
    x = "Parameter f",
    y = "Parameter r",
    fill = "log(BF)"
  )+
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12)
  )


ggsave(
  plot = Plot_logml_heatmap,
  path = figure_path,
  filename = "Plot_M3_OL2019_MC_heatmap_logml.pdf",
  width = 5.5,
  height = 4.5
)

  
```

### 4.4.6 Posterior

```{r}

Model_M3<- read_rds(str_glue("{model_path}Model_{task}_M3_OL2019_choicesoftmax_rboth_fa.rds"))

Post_M3 <- Model_M3 %>% 
  gather_draws(`b_.*`, ndraws=8000, regex = TRUE) %>% 
  separate_wider_delim(.variable, "_", names = c(NA, "par", NA), cols_remove = TRUE) %>%
  filter(par!="b") %>% 
  rename(value = .value) %>% 
  mutate(value = ifelse(str_detect(par,"r|f"),inv_logit_scaled(value), value))

Post_M3 %>% write_csv(str_glue('./figures/figures_Data/Post_{task}_filtering.csv'))

```





